---
title: "RcaT binder pooled screen: DMS-style analysis"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does

We designed 88 candidate binders against the RcaT toxin. Cells carrying the
pooled binder library were grown under 6 conditions (no inducer, toxin-only,
library-only at two IPTG levels, and toxin+library at two IPTG levels), each
in duplicate (A/B). RcaT kills the cell unless it's bound/inhibited, so **a
binder's relative read count after selection is a proxy for how well it
protects the cell** - good binders get relatively enriched when both the
toxin and the binder are induced together.

This mirrors the lab's existing plate-based version of the same assay, and
uses the same three roles for the 6 conditions:

- **no_inducer** (antibiotic only): a viability check, not a comparison
  baseline. A binder that doesn't survive plain antibiotic selection isn't a
  usable data point for anything downstream, so it's dropped here rather
  than showing up later as a meaningless fold-change.
- **library_only** (IPTG only, two levels, no toxin): tells you the *innate*
  toxicity of overexpressing each binder on its own, with no RcaT present -
  reported for context, not used to correct the main result.
- **toxin_only** vs. **toxin_plus_library**: the actual comparison. Does
  inducing the binder rescue growth under toxin challenge, relative to toxin
  alone? The plate-based version of this assay additionally normalises both
  arms against a GFP-only control plasmid; we don't have an equivalent GFP
  control in this sequencing-based setup, so this is the direct,
  un-normalised version of that same ratio (see the caveat in Section 5).

Steps below:

1. Load the counts table produced by `scripts/count_reads_per_binder.py`
   (`results/binder_counts.tsv`) and the sample metadata (`config/samples.tsv`).
2. QC: read depth per sample, how many binders are detected, how evenly the
   pool is represented across binders, and a viability filter that drops
   binders that don't survive no-inducer selection.
3. Normalise within each sample (sequencing depth differs sample to sample).
4. Innate toxicity: library-only vs. no-inducer, as context only.
5. The actual question: toxin+library vs. toxin-only - a quick simple-ratio
   look, then a proper count-based test (DESeq2) that accounts for how much
   read depth actually backs each binder's estimate.
6. Replicate reproducibility (raw frequency, A vs. B).
7. Final ranked hit table, flagging binders with too little read depth to
   trust regardless of what their point estimate says.
8. Compare against the collaborator's published plate-based screen of the
   same binders (Sen2 retron), as an independent sanity check.

```{r libraries}
# DESeq2 (and its Bioconductor dependencies) redefine several dplyr verbs
# (count, desc, slice, rename, first...) - load it BEFORE tidyverse so
# tidyverse's versions win and the rest of this notebook behaves as written.
library(DESeq2)      # count-based hit-calling in Section 5
library(tidyverse)   # readr, dplyr, tidyr, ggplot2, purrr
library(scales)      # nice axis labels
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

# scripts/count_reads_per_binder.py only writes a row when a binder has at
# least one read in a sample, so binder x sample combinations with zero
# reads are simply absent here. Fill those in explicitly as count = 0 -
# a binder legitimately dropping to zero in a condition (e.g. a non-binder
# under toxin+library selection) is exactly the kind of result we want to
# see below, not a row that silently disappears from the analysis.
counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, replicate, ara_pct, iptg_mM),
            by = "sample_id")

# Section 5b (DESeq2) needs both replicates to estimate dispersion, and
# Section 6 compares replicate A against replicate B directly - only
# possible if both actually exist in this run (e.g. not yet true for a
# single-replicate test run before the full experiment). Everything before
# Section 5b works identically either way, since averaging one value is a
# no-op.
has_replicates <- all(c("A", "B") %in% unique(samples$replicate))
if (!has_replicates) {
  message("Only replicate(s) ", paste(unique(samples$replicate), collapse = ", "),
          " present - Section 5b's DESeq2 test and Section 6's A-vs-B consistency check ",
          "will be skipped, and the final hit table will be marked provisional ",
          "(not replicate-validated). Re-run this notebook once both replicates exist.")
}

head(counts_long)
```

Reshape to a wide binder x sample matrix - this is the table students should
picture as "the" dataset from here on (rows = binders, columns = samples).

```{r wide-matrix}
counts_wide <- counts_long %>%
  select(binder_id, sample_id, count) %>%
  pivot_wider(names_from = sample_id, values_from = count, values_fill = 0)

counts_wide
```

## 2. QC: sequencing depth and binder detection

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition, replicate) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

If a sample has much lower depth than the others, everything downstream
(especially rare/low-abundance binders) will be noisier for that sample -
keep this table in mind when interpreting later plots. If you ran
`scripts/count_reads_per_binder.py`, also check the companion
`results/binder_counts_qc.tsv` for how many read pairs were dropped as
unmapped / low-MAPQ / mate-mismatch per sample - a sample with a lot of
`mate_reference_mismatch` hits suggests some binders may share long
identical stretches of sequence and can't always be told apart from a single
read pair.

### Library complexity: how evenly is the pool represented?

This pool is a *competitive* assay: every binder's count is really a share
of a fixed total, so if a handful of binders happen to dominate the
population, everyone else is squeezed into very few reads regardless of
whether they're doing anything interesting biologically. Check that before
trusting any comparison downstream.

```{r library-complexity}
complexity <- counts_long %>%
  group_by(sample_id, condition) %>%
  mutate(sample_total = sum(count)) %>%
  arrange(sample_id, desc(count)) %>%
  mutate(rank = row_number(), cum_share = cumsum(count) / sample_total) %>%
  ungroup()

complexity %>%
  ggplot(aes(x = rank, y = cum_share, group = sample_id, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  labs(x = "Binders, ranked by abundance (most abundant first)",
       y = "Cumulative share of reads",
       title = "Library complexity per sample",
       subtitle = "A steep early climb to the dashed line (90%) means a handful of binders dominate the pool") +
  theme_minimal()

# How many of the panel's binders does it take to reach 90% of a sample's reads?
complexity %>%
  filter(cum_share >= 0.9) %>%
  group_by(sample_id, condition) %>%
  summarise(binders_for_90pct_of_reads = min(rank), .groups = "drop") %>%
  arrange(binders_for_90pct_of_reads)
```

If a small number of binders account for most of a sample's reads, that's
usually a pre-existing skew in the pooled plasmid library (unequal molar
ratios when the individual constructs were pooled, or unequal transformation
efficiency) rather than anything that happened during selection - check
whether the same few binders dominate `no_inducer` (before any selection)
and `toxin_only`/`toxin_plus_library` (after) about equally; if so, it's a
starting-pool problem, not a growth-advantage-under-treatment one. Either
way, it means most of the panel is read-starved and any single-point log2FC
for them will be dominated by sampling noise - keep this in mind in
Section 5b and the low-read-count flag in Section 7.

### Viability filter: drop binders that don't survive no-inducer selection

The plate-based version of this assay uses the no-inducer condition
(antibiotic only, no Ara/IPTG) purely as a check of which plasmids survive at
all - a binder that's already essentially gone before any toxin or induction
step isn't a meaningful data point for anything downstream, and would just
show up later as a huge, meaningless fold-change from a near-empty
denominator. We do the same thing here: drop any binder whose average read
count under `no_inducer` falls below a minimum threshold, before any
normalisation or comparison.

```{r viability-filter}
min_no_inducer_reads <- 10  # tune by eye against the histogram below

no_inducer_counts <- counts_long %>%
  filter(condition == "no_inducer") %>%
  group_by(binder_id) %>%
  summarise(mean_no_inducer_count = mean(count), .groups = "drop")

no_inducer_counts %>%
  ggplot(aes(x = mean_no_inducer_count)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = min_no_inducer_reads, linetype = "dashed", colour = "red") +
  labs(x = "Mean no_inducer read count", y = "Number of binders",
       title = "Choosing the viability threshold",
       subtitle = paste("Dashed line: current threshold =", min_no_inducer_reads)) +
  theme_minimal()

dead_binders <- no_inducer_counts %>%
  filter(mean_no_inducer_count < min_no_inducer_reads) %>%
  pull(binder_id)

if (length(dead_binders) > 0) {
  message(length(dead_binders), " binder(s) dropped as non-viable (mean no_inducer count < ",
          min_no_inducer_reads, "): ", paste(dead_binders, collapse = ", "))
} else {
  message("No binders dropped - all binders survive no-inducer selection above threshold.")
}

counts_long <- counts_long %>% filter(!(binder_id %in% dead_binders))
```

Every plot and table from here on only includes binders that actually grow
under plain antibiotic selection.

## 3. Normalisation

Sequencing depth differs between samples, so raw counts aren't directly
comparable. We convert each binder's count to its **relative frequency
within the sample** (equivalent to counts-per-million, just expressed as a
fraction) - this is the natural normalisation for a competitive pooled assay,
since what we actually care about is what fraction of the surviving
population each binder makes up, not its absolute read count.

A small pseudocount is added before taking logs so that binders with zero
reads in a given sample don't produce `-Inf`.

```{r normalise}
pseudocount <- 1

counts_norm <- counts_long %>%
  group_by(sample_id) %>%
  mutate(
    total_reads = sum(count),
    freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_long$binder_id))
  ) %>%
  ungroup()

counts_norm %>% select(sample_id, binder_id, count, total_reads, freq) %>% head()
```

## 4. Innate toxicity of overexpression (reference only)

We average the two replicates' normalised frequency for each condition
first, then take log2 ratios between conditions. (Section 6 checks the
replicates actually agree with each other before we lean on this average.)

```{r condition-means}
condition_means <- counts_norm %>%
  group_by(condition, binder_id) %>%
  summarise(mean_freq = mean(freq), .groups = "drop")

condition_means %>% head()
```

This isn't part of the actual hit-calling comparison in Section 5 - it's
here purely as context, mirroring what the plate-based assay calls "innate
toxicity of the plasmids": how much does overexpressing each binder cost the
cell on its own, with no toxin present at all (`library_only` vs.
`no_inducer`, at both IPTG levels)?

```{r innate-toxicity}
baseline <- condition_means %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, baseline_freq = mean_freq)

innate_toxicity <- condition_means %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / baseline_freq))

innate_toxicity %>%
  ggplot(aes(x = binder_id, y = log2FC)) +
  geom_col() +
  facet_wrap(~condition, ncol = 1) +
  labs(x = "Binder", y = "log2FC vs. no_inducer",
       title = "Innate toxicity of overexpression (no toxin present)",
       subtitle = "Reference only - not used in the hit-calling comparison below") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

Binders far to the negative side here are costly to overexpress regardless
of RcaT - worth keeping in mind if they also show up near the top of the hit
table below (see the caveat in Section 5).

## 5. The actual question: toxin+library vs. toxin-only

This is the direct comparison the plate-based version of this assay makes
(there, both arms are additionally normalised against a GFP-only control
plasmid to correct for general overexpression burden; we don't have an
equivalent GFP control in this sequencing-based setup, so this is the
un-normalised version of that same ratio). We do this at both IPTG
concentrations.

### 5a. Quick look: simple frequency ratio

```{r key-comparison}
toxin_only_mean <- condition_means %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, toxin_only_freq = mean_freq)

key_comparison <- condition_means %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_mean, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / toxin_only_freq))

key_comparison %>%
  ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs. toxin-only",
       title = "Enrichment of each binder when co-induced with toxin",
       subtitle = "Positive = binder rescues growth (protective) relative to toxin alone") +
  theme_minimal()
```

**Caveat:** without a GFP (or other neutral-overexpression) control, this
comparison can't fully separate RcaT-specific protection from a binder
simply being cheap to overexpress. Cross-check any top hit against the
innate toxicity plot above (a hit that's also very costly to overexpress on
its own is less convincing) and against the 0.1 mM vs. 1 mM dose response
below (a true binder should show *more* protection at higher induction; a
burden artifact generally doesn't).

**A second problem with this simple ratio**: it treats every binder's
estimate as equally trustworthy, which Section 2's library-complexity check
showed isn't true here - most binders are backed by very few reads because
a handful of clones dominate the pool. A binder whose count happens to go
from 3 reads to 8 gets exactly the same kind of log2FC number above as one
whose count goes from 30,000 to 80,000, with nothing here to tell them
apart.

### 5b. A count-based test that knows how much to trust each binder

[DESeq2](https://bioconductor.org/packages/DESeq2/) works directly on raw
counts instead of pre-computed frequencies. It models counts with a negative
binomial distribution, and - since two replicates isn't enough to estimate
noise per binder from scratch - borrows strength across all binders to
learn a shared mean-variance trend. That lets it give a low-count binder an
honestly wide, usually-not-significant result instead of a falsely
confident point estimate, and its size-factor normalisation is built to be
robust to a handful of highly-abundant binders dominating the read total
(exactly Section 2's finding), unlike dividing every count by the sample's
total reads. It also gives a p-value and a multiple-testing-corrected q-value
(`padj`) per binder - useful once you're testing ~80 of them at once.

This needs at least 2 replicates per condition to estimate dispersion, so
it's skipped for a single-replicate run (the simple ratio above is used
instead, marked provisional in the final hit table).

```{r deseq2-hit-calling}
if (has_replicates) {
  deseq_counts <- counts_long %>%
    select(binder_id, sample_id, count) %>%
    pivot_wider(names_from = sample_id, values_from = count, values_fill = 0) %>%
    column_to_rownames("binder_id") %>%
    as.matrix()

  deseq_coldata <- samples %>%
    filter(sample_id %in% colnames(deseq_counts)) %>%
    column_to_rownames("sample_id")
  deseq_coldata <- deseq_coldata[colnames(deseq_counts), , drop = FALSE]
  deseq_coldata$condition <- relevel(factor(deseq_coldata$condition), ref = "toxin_only")

  dds <- DESeqDataSetFromMatrix(countData = deseq_counts, colData = deseq_coldata, design = ~condition)
  # "poscounts" (rather than DESeq2's default) copes with how many binders
  # have a zero count in at least one sample - expected given how skewed
  # this library is - which would otherwise make the default size-factor
  # estimator fail.
  dds <- estimateSizeFactors(dds, type = "poscounts")
  dds <- DESeq(dds)

  extract_contrast <- function(cond) {
    results(dds, contrast = c("condition", cond, "toxin_only")) %>%
      as.data.frame() %>%
      rownames_to_column("binder_id") %>%
      mutate(condition = cond)
  }

  deseq_results <- bind_rows(
    extract_contrast("toxin_plus_library_0.1mM"),
    extract_contrast("toxin_plus_library_1mM")
  ) %>%
    rename(log2FC = log2FoldChange)

  deseq_results %>%
    ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = padj < 0.05)) +
    geom_col() +
    geom_errorbar(aes(ymin = log2FC - lfcSE, ymax = log2FC + lfcSE), width = 0.2, alpha = 0.5) +
    coord_flip() +
    facet_wrap(~condition) +
    scale_fill_manual(values = c(`TRUE` = "steelblue", `FALSE` = "grey70"), na.value = "grey70", name = "padj < 0.05") +
    labs(x = "Binder", y = "DESeq2 log2FC vs. toxin-only",
         title = "Enrichment of each binder when co-induced with toxin (DESeq2)",
         subtitle = "Error bars = standard error - low-count binders naturally get wide bars and rarely reach significance") +
    theme_minimal()
} else {
  message("Skipping DESeq2 - it needs at least 2 replicates per condition to estimate dispersion, ",
          "and only one replicate is present in this run.")
}
```

## 6. Replicate reproducibility

Before trusting either result above, check that replicate A and replicate B
agree on normalised frequency (library complexity/correlation). DESeq2
already folds replicate-to-replicate variation directly into its model in
Section 5b (that's what `lfcSE`/`padj` reflect), so there's no separate
manual A-vs-B consistency check needed for its result - this is a check on
the raw data feeding into it, not a second hit-calling step.

```{r replicate-freq-correlation}
if (has_replicates) {
  rep_wide <- counts_norm %>%
    select(sample_id, binder_id, freq, condition, replicate) %>%
    pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = freq)

  print(
    rep_wide %>%
      ggplot(aes(x = A, y = B)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_x_log10(labels = label_number()) +
      scale_y_log10(labels = label_number()) +
      facet_wrap(~condition) +
      labs(title = "Replicate A vs B: normalised binder frequency", x = "Rep A freq", y = "Rep B freq") +
      theme_minimal()
  )

  rep_wide %>%
    group_by(condition) %>%
    summarise(pearson_r = cor(log10(A), log10(B), use = "complete.obs"), .groups = "drop")
} else {
  message("Skipping - only one replicate present.")
}
```

## 7. Final ranked hit table

With replicates, this is the DESeq2 result from Section 5b, ranked by
`padj`. Without replicates (a single-replicate test run), it falls back to
the simple ratio from Section 5a instead, marked **provisional** - nothing
in that case has been checked for reproducibility or given a proper
uncertainty estimate.

Either way, binders with too little read depth to trust are flagged rather
than dropped - `low_count` binders can still show a big log2FC, but Section
2's library-complexity check means that's exactly the group most likely to
be sampling noise, so treat them with real caution regardless of how they
rank.

```{r final-hits}
low_count_threshold <- 20  # flag (don't exclude) binders averaging fewer reads than this across the comparison - tune by eye against Section 2

if (has_replicates) {
  hit_results <- deseq_results %>%
    mutate(low_count = baseMean < low_count_threshold, provisional = FALSE) %>%
    select(binder_id, condition, log2FC, lfcSE, pvalue, padj, baseMean, low_count, provisional) %>%
    arrange(condition, padj)
} else {
  raw_mean_count <- counts_long %>%
    filter(condition %in% c("toxin_only", "toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
    group_by(binder_id) %>%
    summarise(baseMean = mean(count), .groups = "drop")

  hit_results <- key_comparison %>%
    left_join(raw_mean_count, by = "binder_id") %>%
    mutate(low_count = baseMean < low_count_threshold, provisional = TRUE,
           lfcSE = NA_real_, pvalue = NA_real_, padj = NA_real_) %>%
    select(binder_id, condition, log2FC, lfcSE, pvalue, padj, baseMean, low_count, provisional) %>%
    arrange(condition, desc(log2FC))

  message("Single replicate present - using the simple frequency ratio (Section 5a) instead of ",
          "DESeq2, which needs replicates to estimate dispersion. Treat this table as provisional.")
}

hit_results
```

```{r final-plot}
hit_results %>%
  filter(condition == "toxin_plus_library_1mM") %>%
  slice_max(log2FC, n = 20) %>%
  ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = low_count)) +
  geom_col() +
  { if (has_replicates) geom_errorbar(aes(ymin = log2FC - lfcSE, ymax = log2FC + lfcSE), width = 0.2) } +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), name = "low read count") +
  labs(x = "Binder", y = "log2FC vs. toxin-only (1 mM IPTG)",
       title = paste("Top 20 RcaT binders by protection -",
                      if (has_replicates) "DESeq2" else "simple ratio (provisional - single replicate)")) +
  theme_minimal()
```

```{r save-final-table}
write_csv(hit_results, "../results/final_binder_ranking.csv")
```

## 8. Comparison to the published plate-based screen (Sen2 retron)

Our collaborator ran an equivalent screen of these same binders with a
plate-based assay and shared the results (`data/plate_assay_reference_hits.tsv`),
filtered here to `retron == "Sen2"` (RcaT is the toxin from the Sen2
toxin-antitoxin system used in this project; the file's other retron, Eco9,
is a different system and isn't relevant here). Effect size > 1 means
blocking/protective, matching the sign of a positive `log2FC` in our own
result.

**This isn't an apples-to-apples comparison**: their effect sizes are
calculated against an IPTG-low baseline, not our toxin-only baseline, so
treat this as a rough sanity check on direction and relative ranking, not a
strict validation of magnitude.

```{r load-plate-reference}
plate_reference <- read_tsv("../data/plate_assay_reference_hits.tsv", show_col_types = FALSE) %>%
  filter(retron == "Sen2") %>%
  transmute(genename, plate_pval = pval, plate_log2_efsize = log2(efsize), plate_hit = hit)

# Our binder_id values carry a naming-scheme prefix (`result_...` or
# `rcat-binder-..._result_<n>`) that the collaborator's gene names don't -
# strip it down to the same bare name before joining.
binder_name_map <- tibble(binder_id = unique(hit_results$binder_id)) %>%
  mutate(
    genename = binder_id %>%
      str_remove("^rcat-binder-") %>%
      str_remove("^result_") %>%
      str_replace("_result_", "_")
  )

comparison <- hit_results %>%
  left_join(binder_name_map, by = "binder_id") %>%
  left_join(plate_reference, by = "genename") %>%
  filter(!is.na(plate_log2_efsize))

comparison %>%
  ggplot(aes(x = plate_log2_efsize, y = log2FC, colour = plate_hit, shape = low_count)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey40", linetype = "dashed") +
  facet_wrap(~condition) +
  labs(x = "Published plate assay: log2(effect size), Sen2 (vs. IPTG-low)",
       y = paste("This screen:", if (has_replicates) "DESeq2" else "simple ratio (provisional)", "log2FC (toxin+library vs. toxin-only)"),
       title = "This screen vs. the published plate-based assay",
       colour = "Plate-assay hit call", shape = "Low read count") +
  theme_minimal()

# Does excluding the low-read-count binders (flagged in Section 7) change
# the picture? If most of the disagreement comes from binders Section 2
# already told us are read-starved, that's the library-complexity problem
# showing up here too, not a real disagreement between the two assays.
comparison %>%
  group_by(condition) %>%
  summarise(
    pearson_r_all             = cor(log2FC, plate_log2_efsize, use = "complete.obs"),
    pearson_r_excl_low_count  = cor(log2FC[!low_count], plate_log2_efsize[!low_count], use = "complete.obs"),
    n_binders_matched         = n(),
    n_excl_low_count          = sum(!low_count),
    .groups = "drop"
  )
```

```{r save-comparison-table}
write_csv(comparison, "../results/plate_assay_comparison.csv")
```

A positive correlation here is reassuring, but don't over-read a weak one:
this screen's baseline (toxin-only) and the plate assay's (IPTG-low) aren't
the same condition, both assays have their own noise, and a binder missing
from `plate_reference` above just wasn't in that dataset (check
`n_binders_matched` against the number of viable binders from Section 2 to
see how many actually got compared).

## Where to go from here

* The library-complexity check in Section 2 is worth flagging back to
  whoever built/pooled this plasmid library if it shows a handful of
  binders dominating the pool - it affects any future sequencing run of the
  same pool, not just this analysis, and a better-normalised pool (e.g.
  qPCR-checked equimolar pooling before transformation) would fix it at the
  source rather than needing to be corrected for downstream every time.
* Consider also looking at the 0.1 mM vs 1 mM IPTG comparison directly - a
  true binder should show a dose response (more enrichment at higher
  induction), which is another consistency check independent of replicates.
* No GFP (or other neutral overexpression) control is included in this
  sequencing-based run. If one gets added to a future run, normalising both
  `toxin_only` and `toxin_plus_library` against it before taking the log2FC
  in Section 5 would bring this in line with the plate-based assay's full
  method and remove the caveat above.
