---
title: "RcaT binder pooled screen: DMS-style analysis"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does

We designed 88 candidate binders against the RcaT toxin. Cells carrying the
pooled binder library were grown under 6 conditions (no inducer, toxin-only,
library-only at two IPTG levels, and toxin+library at two IPTG levels), each
in duplicate (A/B). RcaT kills the cell unless it's bound/inhibited, so **a
binder's relative read count after selection is a proxy for how well it
protects the cell** - good binders get relatively enriched when both the
toxin and the binder are induced together.

Steps below:

1. Load the counts table produced by `scripts/count_reads_per_binder.py`
   (`results/binder_counts.tsv`) and the sample metadata (`config/samples.tsv`).
2. QC: read depth per sample, how many binders are actually detected.
3. Normalise within each sample (sequencing depth differs sample to sample).
4. log2 fold-changes: first each induced condition vs. no-inducer baseline
   (sanity checks), then the condition that actually answers the biological
   question - toxin+library vs. toxin-only, i.e. does inducing the binder
   rescue growth under toxin challenge relative to toxin alone?
5. Check that replicates A and B agree.
6. Final ranked hit table.

```{r libraries}
library(tidyverse)   # readr, dplyr, tidyr, ggplot2, purrr
library(scales)      # nice axis labels
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

# scripts/count_reads_per_binder.py only writes a row when a binder has at
# least one read in a sample, so binder x sample combinations with zero
# reads are simply absent here. Fill those in explicitly as count = 0 -
# a binder legitimately dropping to zero in a condition (e.g. a non-binder
# under toxin+library selection) is exactly the kind of result we want to
# see below, not a row that silently disappears from the analysis.
counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, replicate, ara_pct, iptg_mM),
            by = "sample_id")

# Section 5 onward compares replicate A against replicate B directly - only
# possible if both actually exist in this run (e.g. not yet true for a
# single-replicate test run before the full experiment). Everything before
# Section 5 works identically either way, since averaging one value is a
# no-op.
has_replicates <- all(c("A", "B") %in% unique(samples$replicate))
if (!has_replicates) {
  message("Only replicate(s) ", paste(unique(samples$replicate), collapse = ", "),
          " present - Section 5's A-vs-B consistency checks will be skipped, and the ",
          "final hit table will be marked provisional (not replicate-validated). ",
          "Re-run this notebook once both replicates exist.")
}

head(counts_long)
```

Reshape to a wide binder x sample matrix - this is the table students should
picture as "the" dataset from here on (rows = binders, columns = samples).

```{r wide-matrix}
counts_wide <- counts_long %>%
  select(binder_id, sample_id, count) %>%
  pivot_wider(names_from = sample_id, values_from = count, values_fill = 0)

counts_wide
```

## 2. QC: sequencing depth and binder detection

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition, replicate) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

If a sample has much lower depth than the others, everything downstream
(especially rare/low-abundance binders) will be noisier for that sample -
keep this table in mind when interpreting later plots. If you ran
`scripts/count_reads_per_binder.py`, also check the companion
`results/binder_counts_qc.tsv` for how many read pairs were dropped as
unmapped / low-MAPQ / mate-mismatch per sample - a sample with a lot of
`mate_reference_mismatch` hits suggests some binders may share long
identical stretches of sequence and can't always be told apart from a single
read pair.

## 3. Normalisation

Sequencing depth differs between samples, so raw counts aren't directly
comparable. We convert each binder's count to its **relative frequency
within the sample** (equivalent to counts-per-million, just expressed as a
fraction) - this is the natural normalisation for a competitive pooled assay,
since what we actually care about is what fraction of the surviving
population each binder makes up, not its absolute read count.

A small pseudocount is added before taking logs so that binders with zero
reads in a given sample don't produce `-Inf`.

```{r normalise}
pseudocount <- 1

counts_norm <- counts_long %>%
  group_by(sample_id) %>%
  mutate(
    total_reads = sum(count),
    freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_long$binder_id))
  ) %>%
  ungroup()

counts_norm %>% select(sample_id, binder_id, count, total_reads, freq) %>% head()
```

## 4. log2 fold-changes

We average the two replicates' normalised frequency for each condition
first, then take log2 ratios between conditions. (Section 5 checks the
replicates actually agree with each other before we lean on this average.)

```{r condition-means}
condition_means <- counts_norm %>%
  group_by(condition, binder_id) %>%
  summarise(mean_freq = mean(freq), .groups = "drop")

condition_means %>% head()
```

### 4a. Sanity-check comparisons: each induced condition vs. no-inducer

These aren't the final answer, but they're a useful check that induction is
doing *something* (e.g. toxin-only should broadly suppress growth relative
to no-inducer, roughly uniformly across binders since none are expressed yet
- IPTG is 0 in that condition).

```{r lfc-vs-baseline}
baseline <- condition_means %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, baseline_freq = mean_freq)

lfc_vs_baseline <- condition_means %>%
  filter(condition != "no_inducer") %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / baseline_freq))

lfc_vs_baseline %>%
  ggplot(aes(x = binder_id, y = log2FC)) +
  geom_col() +
  facet_wrap(~condition, ncol = 1) +
  labs(x = "Binder", y = "log2FC vs. no_inducer", title = "Each condition vs. no-inducer baseline") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

### 4b. The actual question: toxin+library vs. toxin-only

This isolates the effect of *inducing the binder* while the toxin is
present, which is what should separate good binders (protective, so
enriched) from the rest. We do this at both IPTG concentrations.

```{r lfc-key-comparison}
toxin_only_mean <- condition_means %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, toxin_only_freq = mean_freq)

key_comparison <- condition_means %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_mean, by = "binder_id") %>%
  mutate(log2FC_vs_toxinOnly = log2(mean_freq / toxin_only_freq))

key_comparison %>%
  ggplot(aes(x = reorder(binder_id, log2FC_vs_toxinOnly), y = log2FC_vs_toxinOnly, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs. toxin-only",
       title = "Enrichment of each binder when co-induced with toxin",
       subtitle = "Positive = binder rescues growth (protective) relative to toxin alone") +
  theme_minimal()
```

**Careful: this comparison is confounded.** `toxin_only` has zero IPTG, so no binder
is meaningfully overexpressed there - but `toxin_plus_library` has both toxin *and*
induced overexpression. IPTG-driven overexpression carries its own, RcaT-independent
fitness cost (translation burden, misfolding, etc.), and it varies by binder. So a
binder that's simply cheap to overexpress can look "protective" here even if it does
nothing to RcaT. Section 4c below controls for this.

### 4c. Controlling for overexpression burden: the interaction term

We have a full 2x2 design (toxin x IPTG), so we can isolate the toxin-specific
effect properly instead of guessing at a "better" single baseline. The idea:
measure how much toxin's effect changes when the binder is induced, using each
binder's *own* induced/uninduced pair as its own baseline on both sides:

```
true_protection = log2FC(toxin+lib vs library-only)  -  log2FC(toxin-only vs no-inducer)
                    [toxin's effect WITH the binder induced]   [toxin's effect WITHOUT any binder induced]
```

The first term already nets out each binder's own overexpression burden (both its
numerator and denominator are induced at the same IPTG level). The second term is
the toxin's blanket killing effect at baseline leaky expression - roughly constant
across binders, but computed per-binder here anyway since leakiness itself can
differ. Subtracting them leaves only the part that depends on *both* toxin and
induced binder being present together - i.e. actual RcaT-specific rescue.

```{r lfc-interaction}
library_only_mean <- condition_means %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "library_only_")) %>%
  select(binder_id, iptg_level, library_only_freq = mean_freq)

toxin_plus_library_mean <- condition_means %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "toxin_plus_library_")) %>%
  select(binder_id, iptg_level, condition, toxin_plus_library_freq = mean_freq)

interaction_comparison <- toxin_plus_library_mean %>%
  left_join(library_only_mean, by = c("binder_id", "iptg_level")) %>%
  left_join(toxin_only_mean, by = "binder_id") %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(
    true_protection = log2(toxin_plus_library_freq / library_only_freq) -
      log2(toxin_only_freq / baseline_freq)
  )

interaction_comparison %>%
  ggplot(aes(x = reorder(binder_id, true_protection), y = true_protection, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC(toxin+lib vs lib-only) - log2FC(toxin-only vs no-inducer)",
       title = "Toxin-specific protection, controlling for overexpression burden",
       subtitle = "Positive = this binder neutralises the toxin, beyond what its own overexpression burden explains") +
  theme_minimal()
```

This burden-controlled `true_protection` value is what actually answers the
biological question, so it's what Sections 5-6 check for replicate consistency
and rank below - not the simpler, confounded 4b comparison (which is still
useful as a teaching example of exactly this kind of confound).

## 5. Replicate consistency

Before trusting the log2FC table above, check that replicate A and
replicate B agree - both on raw-ish normalised frequency (library
complexity/correlation) and on the derived log2FC (biological
reproducibility of the actual result we care about).

```{r replicate-freq-correlation}
if (has_replicates) {
  rep_wide <- counts_norm %>%
    select(sample_id, binder_id, freq, condition, replicate) %>%
    pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = freq)

  print(
    rep_wide %>%
      ggplot(aes(x = A, y = B)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_x_log10(labels = label_number()) +
      scale_y_log10(labels = label_number()) +
      facet_wrap(~condition) +
      labs(title = "Replicate A vs B: normalised binder frequency", x = "Rep A freq", y = "Rep B freq") +
      theme_minimal()
  )

  rep_wide %>%
    group_by(condition) %>%
    summarise(pearson_r = cor(log10(A), log10(B), use = "complete.obs"), .groups = "drop")
} else {
  message("Skipping - only one replicate present.")
}
```

Now the same check on the burden-controlled `true_protection` value from 4c
that actually feeds the hit-calling below - compute it **per replicate** (not
averaged) and compare A vs B directly.

```{r replicate-lfc-consistency}
per_rep_freq <- counts_norm %>%
  select(sample_id, binder_id, condition, replicate, freq)

no_inducer_per_rep <- per_rep_freq %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, replicate, no_inducer_freq = freq)

toxin_only_per_rep <- per_rep_freq %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, replicate, toxin_only_freq = freq)

library_only_per_rep <- per_rep_freq %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "library_only_")) %>%
  select(binder_id, replicate, iptg_level, library_only_freq = freq)

lfc_per_rep <- per_rep_freq %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "toxin_plus_library_")) %>%
  left_join(library_only_per_rep, by = c("binder_id", "replicate", "iptg_level")) %>%
  left_join(toxin_only_per_rep, by = c("binder_id", "replicate")) %>%
  left_join(no_inducer_per_rep, by = c("binder_id", "replicate")) %>%
  mutate(log2FC = log2(freq / library_only_freq) - log2(toxin_only_freq / no_inducer_freq))

if (has_replicates) {
  lfc_per_rep_wide <- lfc_per_rep %>%
    pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = log2FC)

  print(
    lfc_per_rep_wide %>%
      ggplot(aes(x = A, y = B)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      facet_wrap(~condition) +
      labs(title = "Replicate A vs B: burden-controlled true_protection (from 4c)",
           x = "Rep A true_protection", y = "Rep B true_protection") +
      theme_minimal()
  )

  lfc_per_rep_wide %>%
    group_by(condition) %>%
    summarise(pearson_r = cor(A, B, use = "complete.obs"), .groups = "drop")
} else {
  message("Skipping - only one replicate present.")
}
```

**A binder is only worth trusting as a hit if its two replicates agree** -
flag anything where A and B disagree in sign or differ by a lot, and treat
those as inconclusive rather than as hits or non-hits. (Skipped entirely if
only one replicate exists - there's nothing to check agreement against yet.)

```{r flag-inconsistent}
consistency_threshold <- 1  # max allowed |A - B| in log2FC units; tune by eye from the plot above

if (has_replicates) {
  lfc_flagged <- lfc_per_rep_wide %>%
    mutate(
      same_direction = sign(A) == sign(B),
      abs_diff       = abs(A - B),
      consistent     = same_direction & abs_diff <= consistency_threshold
    )

  lfc_flagged %>% count(condition, consistent)
} else {
  message("Skipping - only one replicate present.")
}
```

## 6. Final ranked hit table

With both replicates, average their burden-controlled `true_protection`
(from 4c) for consistent binders only, and rank. With a single replicate,
rank directly on that replicate's value instead - but mark the table
**provisional**: nothing here has been checked for reproducibility yet, so
treat it as "worth a closer look", not "confirmed hits", until replicate B
exists.

```{r final-hits}
if (has_replicates) {
  final_hits <- lfc_flagged %>%
    filter(consistent) %>%
    mutate(mean_log2FC = (A + B) / 2, provisional = FALSE) %>%
    select(binder_id, condition, mean_log2FC, A, B, provisional) %>%
    arrange(condition, desc(mean_log2FC))
} else {
  final_hits <- lfc_per_rep %>%
    mutate(mean_log2FC = log2FC, A = log2FC, B = NA_real_, provisional = TRUE) %>%
    select(binder_id, condition, mean_log2FC, A, B, provisional) %>%
    arrange(condition, desc(mean_log2FC))
}

final_hits
```

```{r final-plot}
final_hits %>%
  filter(condition == "toxin_plus_library_1mM") %>%
  slice_max(mean_log2FC, n = 20) %>%
  ggplot(aes(x = reorder(binder_id, mean_log2FC), y = mean_log2FC, fill = provisional)) +
  geom_col() +
  { if (has_replicates) geom_errorbar(aes(ymin = pmin(A, B), ymax = pmax(A, B)), width = 0.2) } +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), guide = if (has_replicates) "none" else "legend") +
  labs(x = "Binder", y = "true_protection (1 mM IPTG, burden-controlled)",
       title = paste("Top 20 RcaT binders by", if (has_replicates) "replicate-consistent, burden-controlled protection" else "burden-controlled protection (provisional - single replicate)")) +
  theme_minimal()
```

```{r save-final-table}
write_csv(final_hits, "../results/final_binder_ranking.csv")
```

## Where to go from here

* This uses a simple frequency-ratio log2FC, which is easy to reason about
  but doesn't model count-based variance the way tools like DESeq2/edgeR do
  (which down-weight noisy, low-count binders and give you a p-value per
  binder). If you want to go further, feeding `counts_wide` into DESeq2 with
  a design of `~ condition` and pulling the `toxin_plus_library_*` vs.
  `toxin_only` contrast is a natural next step.
* Consider also looking at the 0.1 mM vs 1 mM IPTG comparison directly - a
  true binder should show a dose response (more enrichment at higher
  induction), which is another consistency check independent of replicates.
