---
title: "RcaT binder pooled screen: DMS-style analysis (high/low-expression batches)"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does differently

This is a variant of `DMS_analysis.Rmd` (see that notebook for the full
walkthrough of the experiment, the conditions, and the basic logic - it's
not repeated here). The difference is in how it handles something that
notebook's library-complexity check (Section 2) turned up: in this pool, a
handful of binders dominate the read counts in every sample, while most of
the other ~80 are read-starved by comparison. Because every binder's count
is really a *share* of a fixed sequencing total, those dominant binders
distort the normalisation (and DESeq2's size factors) for everyone else -
even a binder whose own abundance didn't really change can look enriched or
depleted just because the dominant clones' share shifted around it.

**The fix here**: split the panel into a high-expressing batch and a
low-expressing batch *before* any normalisation or hit-calling, and run the
entire comparison (simple ratio, then DESeq2) separately within each batch.
Low-expressing binders are never thrown out - they're just compared only
against other binders at a similar abundance, instead of being
compositionally squeezed by a few dominant clones that have nothing to do
with them.

**What this does and doesn't fix**: splitting removes the dominant clones
from the low-expressing batch's own normalisation, which is a real
distortion. It does *not* manufacture extra sequencing depth - a binder with
genuinely few reads still has genuinely few reads, and DESeq2 will still
(correctly) flag it as low-confidence. Splitting fixes a normalisation bias,
not a sequencing-depth limit.

```{r libraries}
# DESeq2 (and its Bioconductor dependencies) redefine several dplyr verbs
# (count, desc, slice, rename, first...) - load it BEFORE tidyverse so
# tidyverse's versions win and the rest of this notebook behaves as written.
library(DESeq2)
library(tidyverse)
library(scales)
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, replicate, ara_pct, iptg_mM),
            by = "sample_id")

# DESeq2 needs at least 2 replicates per condition to estimate dispersion;
# everything before the batch pipeline works identically either way.
has_replicates <- all(c("A", "B") %in% unique(samples$replicate))
if (!has_replicates) {
  message("Only replicate(s) ", paste(unique(samples$replicate), collapse = ", "),
          " present - DESeq2 will be skipped in both batches below, and both ",
          "final hit tables will be marked provisional.")
}

head(counts_long)
```

## 2. QC: sequencing depth, library complexity, viability

Same QC as `DMS_analysis.Rmd` - included here so the batch split below has
a clean, already-filtered starting point.

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition, replicate) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

### Library complexity: how evenly is the pool represented?

```{r library-complexity}
complexity <- counts_long %>%
  group_by(sample_id, condition) %>%
  mutate(sample_total = sum(count)) %>%
  arrange(sample_id, desc(count)) %>%
  mutate(rank = row_number(), cum_share = cumsum(count) / sample_total) %>%
  ungroup()

complexity %>%
  ggplot(aes(x = rank, y = cum_share, group = sample_id, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  labs(x = "Binders, ranked by abundance (most abundant first)",
       y = "Cumulative share of reads",
       title = "Library complexity per sample",
       subtitle = "A steep early climb to the dashed line (90%) means a handful of binders dominate the pool") +
  theme_minimal()
```

This is exactly the pattern that motivates the batch split below - if it's
flat/gradual here (no dominant clones), splitting into batches won't change
much and you may as well use the unsplit `DMS_analysis.Rmd` instead.

### Viability filter: drop binders that don't survive no-inducer selection

```{r viability-filter}
min_no_inducer_reads <- 10  # tune by eye against the histogram below

no_inducer_counts <- counts_long %>%
  filter(condition == "no_inducer") %>%
  group_by(binder_id) %>%
  summarise(mean_no_inducer_count = mean(count), .groups = "drop")

no_inducer_counts %>%
  ggplot(aes(x = mean_no_inducer_count)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = min_no_inducer_reads, linetype = "dashed", colour = "red") +
  labs(x = "Mean no_inducer read count", y = "Number of binders",
       title = "Choosing the viability threshold",
       subtitle = paste("Dashed line: current threshold =", min_no_inducer_reads)) +
  theme_minimal()

dead_binders <- no_inducer_counts %>%
  filter(mean_no_inducer_count < min_no_inducer_reads) %>%
  pull(binder_id)

if (length(dead_binders) > 0) {
  message(length(dead_binders), " binder(s) dropped as non-viable (mean no_inducer count < ",
          min_no_inducer_reads, "): ", paste(dead_binders, collapse = ", "))
} else {
  message("No binders dropped - all binders survive no-inducer selection above threshold.")
}

counts_long <- counts_long %>% filter(!(binder_id %in% dead_binders))
```

### Splitting into high- and low-expressing batches

Rank binders by their overall abundance (mean raw read count across *all*
samples and conditions - not just one condition, since Section 2's library
complexity check showed the same handful of binders dominate every
condition about equally). Binders at or above the threshold go in the
high-expressing batch; everyone else goes in the low-expressing batch.
Nothing is excluded here - every viable binder lands in exactly one batch.

```{r expression-batch-split}
high_expr_threshold <- 1000  # mean read count across all samples; tune by eye against the histogram below

overall_abundance <- counts_long %>%
  group_by(binder_id) %>%
  summarise(mean_count_overall = mean(count), .groups = "drop")

overall_abundance %>%
  ggplot(aes(x = mean_count_overall)) +
  geom_histogram(bins = 30) +
  scale_x_log10(labels = label_number()) +
  geom_vline(xintercept = high_expr_threshold, linetype = "dashed", colour = "red") +
  labs(x = "Mean read count across all samples (log scale)", y = "Number of binders",
       title = "Choosing the high/low expression split",
       subtitle = paste("Dashed line: current threshold =", high_expr_threshold)) +
  theme_minimal()

expr_batch <- overall_abundance %>%
  mutate(batch = if_else(mean_count_overall >= high_expr_threshold, "high", "low"))

high_expr_binders <- expr_batch %>% filter(batch == "high") %>% pull(binder_id)
low_expr_binders  <- expr_batch %>% filter(batch == "low") %>% pull(binder_id)

message(length(high_expr_binders), " binder(s) assigned to the high-expressing batch, ",
        length(low_expr_binders), " to the low-expressing batch.")

# Sanity check: the high batch should hold most of the READS; the low batch
# should hold most of the BINDERS. That mismatch is the whole reason to split.
counts_long %>%
  left_join(expr_batch %>% select(binder_id, batch), by = "binder_id") %>%
  group_by(batch) %>%
  summarise(n_binders = n_distinct(binder_id), total_reads = sum(count), .groups = "drop") %>%
  mutate(share_of_reads = total_reads / sum(total_reads))
```

## 3. The per-batch pipeline

Everything from here on (normalisation, innate toxicity, the simple-ratio
comparison, DESeq2, the final hit table) is identical in logic to
`DMS_analysis.Rmd` Sections 3-7 - it's written once as a function and run
twice (Sections 4 and 5), so the two batches get exactly the same treatment
and nothing diverges by accident. The key point: every `group_by(sample_id)`
normalisation step and DESeq2's size-factor estimation only ever see the
binders in *that* batch - a low-expressing binder's frequency is computed
against the low batch's own total, never diluted by the high batch's reads.

```{r batch-pipeline-functions}
pseudocount <- 1
low_count_threshold <- 20  # flag (don't exclude) binders averaging fewer reads than this - tune by eye

run_batch_pipeline <- function(counts_batch, samples, has_replicates) {
  counts_norm <- counts_batch %>%
    group_by(sample_id) %>%
    mutate(
      total_reads = sum(count),
      freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_batch$binder_id))
    ) %>%
    ungroup()

  condition_means <- counts_norm %>%
    group_by(condition, binder_id) %>%
    summarise(mean_freq = mean(freq), .groups = "drop")

  baseline <- condition_means %>%
    filter(condition == "no_inducer") %>%
    select(binder_id, baseline_freq = mean_freq)

  innate_toxicity <- condition_means %>%
    filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
    left_join(baseline, by = "binder_id") %>%
    mutate(log2FC = log2(mean_freq / baseline_freq))

  p_innate_toxicity <- innate_toxicity %>%
    ggplot(aes(x = binder_id, y = log2FC)) +
    geom_col() +
    facet_wrap(~condition, ncol = 1) +
    labs(x = "Binder", y = "log2FC vs. no_inducer",
         title = "Innate toxicity of overexpression (no toxin present)",
         subtitle = "Reference only - not used in hit-calling") +
    theme_minimal() +
    theme(axis.text.x = element_blank())

  # 5a. Quick look: simple frequency ratio, normalised within this batch only
  toxin_only_mean <- condition_means %>%
    filter(condition == "toxin_only") %>%
    select(binder_id, toxin_only_freq = mean_freq)

  key_comparison <- condition_means %>%
    filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
    left_join(toxin_only_mean, by = "binder_id") %>%
    mutate(log2FC = log2(mean_freq / toxin_only_freq))

  p_key_comparison <- key_comparison %>%
    ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = condition)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(x = "Binder", y = "log2FC vs. toxin-only",
         title = "Enrichment when co-induced with toxin (simple ratio, this batch only)",
         subtitle = "Positive = binder rescues growth (protective) relative to toxin alone") +
    theme_minimal()

  # 5b. DESeq2, fit on this batch's count matrix only - its own size factors
  # and dispersion trend, untouched by the other batch's binders
  deseq_results <- NULL
  p_deseq <- NULL
  if (has_replicates) {
    deseq_counts <- counts_batch %>%
      select(binder_id, sample_id, count) %>%
      pivot_wider(names_from = sample_id, values_from = count, values_fill = 0) %>%
      column_to_rownames("binder_id") %>%
      as.matrix()

    deseq_coldata <- samples %>%
      filter(sample_id %in% colnames(deseq_counts)) %>%
      column_to_rownames("sample_id")
    deseq_coldata <- deseq_coldata[colnames(deseq_counts), , drop = FALSE]
    deseq_coldata$condition <- relevel(factor(deseq_coldata$condition), ref = "toxin_only")

    dds <- DESeqDataSetFromMatrix(countData = deseq_counts, colData = deseq_coldata, design = ~condition)
    dds <- estimateSizeFactors(dds, type = "poscounts")
    # A batch can be just a handful of binders (the high-expressing one
    # especially), too few for DESeq2's usual trend-curve dispersion fit -
    # it errors rather than silently doing something unreasonable. When
    # that happens, fall back to each binder's own gene-wise dispersion
    # estimate directly (no shrinkage toward a trend), which is DESeq2's
    # own documented recommendation for a dataset with very few features.
    dds <- tryCatch(
      DESeq(dds),
      error = function(e) {
        message("Trend-curve dispersion fit failed (too few/too-similar binders in this batch) - ",
                "falling back to gene-wise dispersion estimates.")
        d <- estimateDispersionsGeneEst(dds)
        dispersions(d) <- S4Vectors::mcols(d)$dispGeneEst
        nbinomWaldTest(d)
      }
    )

    extract_contrast <- function(cond) {
      results(dds, contrast = c("condition", cond, "toxin_only")) %>%
        as.data.frame() %>%
        rownames_to_column("binder_id") %>%
        mutate(condition = cond)
    }

    deseq_results <- bind_rows(
      extract_contrast("toxin_plus_library_0.1mM"),
      extract_contrast("toxin_plus_library_1mM")
    ) %>%
      rename(log2FC = log2FoldChange)

    p_deseq <- deseq_results %>%
      ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = padj < 0.05)) +
      geom_col() +
      geom_errorbar(aes(ymin = log2FC - lfcSE, ymax = log2FC + lfcSE), width = 0.2, alpha = 0.5) +
      coord_flip() +
      facet_wrap(~condition) +
      scale_fill_manual(values = c(`TRUE` = "steelblue", `FALSE` = "grey70"), na.value = "grey70", name = "padj < 0.05") +
      labs(x = "Binder", y = "DESeq2 log2FC vs. toxin-only",
           title = "Enrichment when co-induced with toxin (DESeq2, this batch only)",
           subtitle = "Error bars = standard error") +
      theme_minimal()
  }

  # 7. Final hit table for this batch
  if (has_replicates) {
    hit_results <- deseq_results %>%
      mutate(low_count = baseMean < low_count_threshold, provisional = FALSE) %>%
      select(binder_id, condition, log2FC, lfcSE, pvalue, padj, baseMean, low_count, provisional) %>%
      arrange(condition, padj)
  } else {
    raw_mean_count <- counts_batch %>%
      filter(condition %in% c("toxin_only", "toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
      group_by(binder_id) %>%
      summarise(baseMean = mean(count), .groups = "drop")

    hit_results <- key_comparison %>%
      left_join(raw_mean_count, by = "binder_id") %>%
      mutate(low_count = baseMean < low_count_threshold, provisional = TRUE,
             lfcSE = NA_real_, pvalue = NA_real_, padj = NA_real_) %>%
      select(binder_id, condition, log2FC, lfcSE, pvalue, padj, baseMean, low_count, provisional) %>%
      arrange(condition, desc(log2FC))
  }

  p_final <- hit_results %>%
    filter(condition == "toxin_plus_library_1mM") %>%
    slice_max(log2FC, n = 20) %>%
    ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = low_count)) +
    geom_col() +
    { if (has_replicates) geom_errorbar(aes(ymin = log2FC - lfcSE, ymax = log2FC + lfcSE), width = 0.2) } +
    coord_flip() +
    scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), name = "low read count") +
    labs(x = "Binder", y = "log2FC vs. toxin-only (1 mM IPTG)",
         title = paste("Top binders by protection -", if (has_replicates) "DESeq2" else "simple ratio (provisional)")) +
    theme_minimal()

  list(
    innate_toxicity = innate_toxicity, p_innate_toxicity = p_innate_toxicity,
    key_comparison = key_comparison, p_key_comparison = p_key_comparison,
    deseq_results = deseq_results, p_deseq = p_deseq,
    hit_results = hit_results, p_final = p_final
  )
}

compare_to_plate <- function(hit_results, plate_reference, batch_label) {
  binder_name_map <- tibble(binder_id = unique(hit_results$binder_id)) %>%
    mutate(
      genename = binder_id %>%
        str_remove("^rcat-binder-") %>%
        str_remove("^result_") %>%
        str_replace("_result_", "_")
    )

  comparison <- hit_results %>%
    left_join(binder_name_map, by = "binder_id") %>%
    left_join(plate_reference, by = "genename") %>%
    filter(!is.na(plate_log2_efsize))

  p <- comparison %>%
    ggplot(aes(x = plate_log2_efsize, y = log2FC, colour = plate_hit, shape = low_count)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linetype = "dashed") +
    facet_wrap(~condition) +
    labs(x = "Published plate assay: log2(effect size), Sen2 (vs. IPTG-low)",
         y = paste0(batch_label, "-batch log2FC (toxin+library vs. toxin-only)"),
         title = paste0("This screen (", batch_label, " batch) vs. the published plate-based assay"),
         colour = "Plate-assay hit call", shape = "Low read count") +
    theme_minimal()

  cor_summary <- comparison %>%
    group_by(condition) %>%
    summarise(
      pearson_r_all            = cor(log2FC, plate_log2_efsize, use = "complete.obs"),
      pearson_r_excl_low_count = cor(log2FC[!low_count], plate_log2_efsize[!low_count], use = "complete.obs"),
      n_binders_matched        = n(),
      n_excl_low_count         = sum(!low_count),
      .groups = "drop"
    ) %>%
    mutate(batch = batch_label, .before = 1)

  list(comparison = comparison, plot = p, cor_summary = cor_summary)
}
```

## 4. High-expressing batch

```{r high-batch-run}
high_batch_counts <- counts_long %>% filter(binder_id %in% high_expr_binders)
high <- run_batch_pipeline(high_batch_counts, samples, has_replicates)
```

```{r high-innate-toxicity}
high$p_innate_toxicity
```

```{r high-key-comparison}
high$p_key_comparison
```

```{r high-deseq2}
if (has_replicates) {
  print(high$p_deseq)
} else {
  message("Skipping DESeq2 - only one replicate present.")
}
```

**Caution**: this batch is, by construction, a small handful of binders -
DESeq2's dispersion estimation (and any correlation computed from it below)
is far less reliable with this few data points than it is for the
low-expressing batch.

```{r high-final-hits}
high$hit_results
```

```{r high-final-plot}
high$p_final
```

```{r high-save}
write_csv(high$hit_results, "../results/final_binder_ranking_high.csv")
```

## 5. Low-expressing batch

This is the batch that was getting compositionally squeezed in the unsplit
analysis - here it's normalised and DESeq2-tested entirely on its own terms.

```{r low-batch-run}
low_batch_counts <- counts_long %>% filter(binder_id %in% low_expr_binders)
low <- run_batch_pipeline(low_batch_counts, samples, has_replicates)
```

```{r low-innate-toxicity}
low$p_innate_toxicity
```

```{r low-key-comparison}
low$p_key_comparison
```

```{r low-deseq2}
if (has_replicates) {
  print(low$p_deseq)
} else {
  message("Skipping DESeq2 - only one replicate present.")
}
```

```{r low-final-hits}
low$hit_results
```

```{r low-final-plot}
low$p_final
```

```{r low-save}
write_csv(low$hit_results, "../results/final_binder_ranking_low.csv")
```

## 6. Comparing each batch to the published plate-based screen (Sen2 retron)

Same reference data and same caveats as `DMS_analysis.Rmd` Section 8
(their effect size uses an IPTG-low baseline, not our toxin-only baseline -
this is a rough sanity check, not a strict validation) - run once per batch.

```{r load-plate-reference}
plate_reference <- read_tsv("../data/plate_assay_reference_hits.tsv", show_col_types = FALSE) %>%
  filter(retron == "Sen2") %>%
  transmute(genename, plate_pval = pval, plate_log2_efsize = log2(efsize), plate_hit = hit)
```

```{r plate-compare-batches}
cmp_high <- compare_to_plate(high$hit_results, plate_reference, "high-expressing")
cmp_low  <- compare_to_plate(low$hit_results,  plate_reference, "low-expressing")

print(cmp_high$plot)
print(cmp_low$plot)

bind_rows(cmp_high$cor_summary, cmp_low$cor_summary)
```

```{r save-plate-comparisons}
write_csv(cmp_high$comparison, "../results/plate_assay_comparison_high.csv")
write_csv(cmp_low$comparison, "../results/plate_assay_comparison_low.csv")
```

## 7. Did splitting help?

Compare the table above against `DMS_analysis.Rmd`'s unsplit
`plate_assay_comparison.csv` correlation. What to look for:

- If the low-expressing batch's correlation with the plate assay is
  noticeably better here than in the unsplit notebook, that's evidence the
  dominant clones really were distorting the comparison for everyone else -
  exactly the problem this notebook exists to fix.
- If it's about the same, the low-expressing binders' own read depth (not
  the dominant clones' compositional pull) is the limiting factor - check
  `n_excl_low_count` and how many of the low batch are flagged `low_count`;
  splitting can't fix a binder that was only ever backed by a handful of
  reads to begin with.
- The high-expressing batch's correlation is based on very few binders
  (see the caution in Section 4) - a single outlier among 5-10 points can
  swing that number a lot, so don't read much into it alone.

## Where to go from here

* If most binders land in one batch and almost none in the other, the
  `high_expr_threshold` in Section 2 is in the wrong place for this
  dataset - move it and re-knit; the histogram there is the thing to look
  at, not a fixed number.
* This splits on overall abundance once, globally. A binder that's
  middling in one condition but dominant in another would still be
  assigned to a single batch for every comparison - if that matters for
  your data, consider a per-condition rather than per-binder split.
* Everything in `DMS_analysis.Rmd`'s own "Where to go from here" still
  applies here too (DESeq2 vs. simple ratio trade-offs, the dose-response
  check, the missing GFP control).
