---
title: "RcaT binder pooled screen: DMS-style analysis"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does

We designed 88 candidate binders against the RcaT toxin. Cells carrying the
pooled binder library were grown under 6 conditions (no inducer, toxin-only,
library-only at two IPTG levels, and toxin+library at two IPTG levels), one
sample per condition (no biological replicates - see the note below). RcaT
kills the cell unless it's bound/inhibited, so **a binder's relative read
count after selection is a proxy for how well it protects the cell** - good
binders get relatively enriched when both the toxin and the binder are
induced together.

This mirrors the lab's existing plate-based version of the same assay, and
uses the same three roles for the 6 conditions:

- **no_inducer** (antibiotic only): a viability check, not a comparison
  baseline. A binder that doesn't survive plain antibiotic selection isn't a
  usable data point for anything downstream, so it's dropped here rather
  than showing up later as a meaningless fold-change.
- **library_only** (IPTG only, two levels, no toxin): tells you the *innate*
  toxicity of overexpressing each binder on its own, with no RcaT present -
  reported for context in Section 4, and also used as the baseline for the
  main comparison below.
- **toxin_plus_library vs. library_only, matched IPTG level**: the actual
  comparison. Both arms have the binder induced at the same level, so each
  binder's own overexpression cost is present on both sides of the ratio and
  should mostly cancel out - what's left over is specifically the effect of
  adding the toxin on top of an already-expressed binder (see Section 5).

**No replicates**: this run has a single sample per condition. Every value
below is a one-off point estimate, not an average of replicates - there's no
A-vs-B consistency check to lean on, and no count-based statistical test
(DESeq2 and similar tools need at least 2 replicates per condition to
estimate dispersion, so none is used here). Section 6's low-read-count flag
is the main safeguard against over-trusting a noisy point estimate.

Steps below:

1. Load the counts table produced by `scripts/count_reads_per_binder.py`
   (`results/binder_counts.tsv`) and the sample metadata (`config/samples.tsv`).
2. QC: read depth per sample, how many binders are detected, how evenly the
   pool is represented across binders, and a viability filter that drops
   binders that don't survive no-inducer selection.
3. Normalise within each sample (sequencing depth differs sample to sample).
4. Innate toxicity: library-only vs. no-inducer, as context only.
5. The actual question: toxin+library vs. library-only, at matched IPTG
   levels.
6. Final ranked hit table, flagging binders with too little read depth to
   trust regardless of what their point estimate says.
7. Compare against the collaborator's published plate-based screen of the
   same binders (Sen2 retron), as an independent sanity check.

```{r libraries}
library(tidyverse)   # readr, dplyr, tidyr, ggplot2, purrr
library(scales)      # nice axis labels
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

# scripts/count_reads_per_binder.py only writes a row when a binder has at
# least one read in a sample, so binder x sample combinations with zero
# reads are simply absent here. Fill those in explicitly as count = 0 -
# a binder legitimately dropping to zero in a condition (e.g. a non-binder
# under toxin+library selection) is exactly the kind of result we want to
# see below, not a row that silently disappears from the analysis.
counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, ara_pct, iptg_mM),
            by = "sample_id")

head(counts_long)
```

Reshape to a wide binder x sample matrix - this is the table students should
picture as "the" dataset from here on (rows = binders, columns = samples).

```{r wide-matrix}
counts_wide <- counts_long %>%
  select(binder_id, sample_id, count) %>%
  pivot_wider(names_from = sample_id, values_from = count, values_fill = 0)

counts_wide
```

## 2. QC: sequencing depth and binder detection

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

If a sample has much lower depth than the others, everything downstream
(especially rare/low-abundance binders) will be noisier for that sample -
keep this table in mind when interpreting later plots. If you ran
`scripts/count_reads_per_binder.py`, also check the companion
`results/binder_counts_qc.tsv` for how many read pairs were dropped as
unmapped / low-MAPQ / mate-mismatch per sample - a sample with a lot of
`mate_reference_mismatch` hits suggests some binders may share long
identical stretches of sequence and can't always be told apart from a single
read pair.

### Library complexity: how evenly is the pool represented?

This pool is a *competitive* assay: every binder's count is really a share
of a fixed total, so if a handful of binders happen to dominate the
population, everyone else is squeezed into very few reads regardless of
whether they're doing anything interesting biologically. Check that before
trusting any comparison downstream.

```{r library-complexity}
complexity <- counts_long %>%
  group_by(sample_id, condition) %>%
  mutate(sample_total = sum(count)) %>%
  arrange(sample_id, desc(count)) %>%
  mutate(rank = row_number(), cum_share = cumsum(count) / sample_total) %>%
  ungroup()

complexity %>%
  ggplot(aes(x = rank, y = cum_share, group = sample_id, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  labs(x = "Binders, ranked by abundance (most abundant first)",
       y = "Cumulative share of reads",
       title = "Library complexity per sample",
       subtitle = "A steep early climb to the dashed line (90%) means a handful of binders dominate the pool") +
  theme_minimal()

# How many of the panel's binders does it take to reach 90% of a sample's reads?
complexity %>%
  filter(cum_share >= 0.9) %>%
  group_by(sample_id, condition) %>%
  summarise(binders_for_90pct_of_reads = min(rank), .groups = "drop") %>%
  arrange(binders_for_90pct_of_reads)
```

If a small number of binders account for most of a sample's reads, that's
usually a pre-existing skew in the pooled plasmid library (unequal molar
ratios when the individual constructs were pooled, or unequal transformation
efficiency) rather than anything that happened during selection - check
whether the same few binders dominate `no_inducer` (before any selection)
and `toxin_only`/`toxin_plus_library` (after) about equally; if so, it's a
starting-pool problem, not a growth-advantage-under-treatment one. Either
way, it means most of the panel is read-starved and any single-point log2FC
for them will be dominated by sampling noise - keep this in mind in
Section 5 and the low-read-count flag in Section 6.

### Viability filter: drop binders that don't survive no-inducer selection

The plate-based version of this assay uses the no-inducer condition
(antibiotic only, no Ara/IPTG) purely as a check of which plasmids survive at
all - a binder that's already essentially gone before any toxin or induction
step isn't a meaningful data point for anything downstream, and would just
show up later as a huge, meaningless fold-change from a near-empty
denominator. We do the same thing here: drop any binder whose read count
under `no_inducer` falls below a minimum threshold, before any normalisation
or comparison.

```{r viability-filter}
min_no_inducer_reads <- 10  # tune by eye against the histogram below

no_inducer_counts <- counts_long %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, no_inducer_count = count)

no_inducer_counts %>%
  ggplot(aes(x = no_inducer_count)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = min_no_inducer_reads, linetype = "dashed", colour = "red") +
  labs(x = "no_inducer read count", y = "Number of binders",
       title = "Choosing the viability threshold",
       subtitle = paste("Dashed line: current threshold =", min_no_inducer_reads)) +
  theme_minimal()

dead_binders <- no_inducer_counts %>%
  filter(no_inducer_count < min_no_inducer_reads) %>%
  pull(binder_id)

if (length(dead_binders) > 0) {
  message(length(dead_binders), " binder(s) dropped as non-viable (no_inducer count < ",
          min_no_inducer_reads, "): ", paste(dead_binders, collapse = ", "))
} else {
  message("No binders dropped - all binders survive no-inducer selection above threshold.")
}

counts_long <- counts_long %>% filter(!(binder_id %in% dead_binders))
```

Every plot and table from here on only includes binders that actually grow
under plain antibiotic selection.

## 3. Normalisation

Sequencing depth differs between samples, so raw counts aren't directly
comparable. We convert each binder's count to its **relative frequency
within the sample** (equivalent to counts-per-million, just expressed as a
fraction) - this is the natural normalisation for a competitive pooled assay,
since what we actually care about is what fraction of the surviving
population each binder makes up, not its absolute read count.

A small pseudocount is added before taking logs so that binders with zero
reads in a given sample don't produce `-Inf`.

```{r normalise}
pseudocount <- 1

counts_norm <- counts_long %>%
  group_by(sample_id) %>%
  mutate(
    total_reads = sum(count),
    freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_long$binder_id))
  ) %>%
  ungroup()

counts_norm %>% select(sample_id, binder_id, count, total_reads, freq) %>% head()
```

## 4. Innate toxicity of overexpression (reference only)

```{r condition-freq}
condition_freq <- counts_norm %>%
  select(condition, binder_id, freq)

condition_freq %>% head()
```

This isn't part of the actual hit-calling comparison in Section 5 - it's
here purely as context, mirroring what the plate-based assay calls "innate
toxicity of the plasmids": how much does overexpressing each binder cost the
cell on its own, with no toxin present at all (`library_only` vs.
`no_inducer`, at both IPTG levels)?

```{r innate-toxicity}
baseline <- condition_freq %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, baseline_freq = freq)

innate_toxicity <- condition_freq %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(log2FC = log2(freq / baseline_freq))

innate_toxicity %>%
  ggplot(aes(x = binder_id, y = log2FC)) +
  geom_col() +
  facet_wrap(~condition, ncol = 1) +
  labs(x = "Binder", y = "log2FC vs. no_inducer",
       title = "Innate toxicity of overexpression (no toxin present)",
       subtitle = "Reference only - not used in the hit-calling comparison below") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

Binders far to the negative side here are costly to overexpress regardless
of RcaT - worth keeping in mind if they also show up near the top of the hit
table below (see the caveat in Section 5).

## 5. The actual question: toxin+library vs. library-only (matched IPTG level)

Both arms here have the binder induced at the same IPTG level, so each
binder's own overexpression cost is present on both sides of the ratio and
should mostly cancel out - what's left over is mainly the effect of adding
the toxin on top of an already-expressed binder. (An earlier version of this
notebook compared against `toxin_only` instead - with no IPTG at all there,
that comparison ended up dominated by each binder's own innate overexpression
cost, which has nothing to do with RcaT. Matching the IPTG level removes
most of that confound.)

```{r key-comparison}
library_only_freq <- condition_freq %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "library_only_")) %>%
  select(binder_id, iptg_level, library_only_freq = freq)

key_comparison <- condition_freq %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  mutate(iptg_level = str_remove(condition, "toxin_plus_library_")) %>%
  left_join(library_only_freq, by = c("binder_id", "iptg_level")) %>%
  mutate(log2FC = log2(freq / library_only_freq)) %>%
  select(-iptg_level)

key_comparison %>%
  ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs. library-only (same IPTG level)",
       title = "Enrichment of each binder when co-induced with toxin",
       subtitle = "Positive = binder does better under toxin challenge than its own baseline without toxin") +
  theme_minimal()
```

**Caveat:** this assumes each binder's own overexpression cost is about the
same with or without toxin present - a reasonable first approximation, but
not guaranteed (the extra stress of active RcaT could itself change how
costly a binder is to express). It's also not a substitute for a true GFP
(neutral-overexpression) control, which the plate-based version of this
assay uses and we don't have here. Cross-check any top hit against the
innate toxicity plot above (a hit that's also very costly to overexpress on
its own is less convincing) and against the 0.1 mM vs. 1 mM dose response
(a true binder should show *more* protection at higher induction; a burden
artifact generally doesn't).

**With a single sample per condition**, every value above is a one-off point
estimate - there's no replicate to check it against, and no count-based
statistical test (DESeq2 and similar tools need at least 2 replicates per
condition) to give it a proper uncertainty estimate. Section 6's
low-read-count flag is the main safeguard against over-trusting a noisy
point estimate; it doesn't replace having a second replicate.

## 6. Final ranked hit table

Binders with too little read depth to trust are flagged rather than
dropped - `low_count` binders can still show a big log2FC, but Section 2's
library-complexity check means that's exactly the group most likely to be
sampling noise, so treat them with real caution regardless of how they rank.

```{r final-hits}
low_count_threshold <- 20  # flag (don't exclude) binders averaging fewer reads than this across the compared conditions - tune by eye against Section 2

raw_mean_count <- counts_long %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM",
                           "toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  group_by(binder_id) %>%
  summarise(baseMean = mean(count), .groups = "drop")

hit_results <- key_comparison %>%
  left_join(raw_mean_count, by = "binder_id") %>%
  mutate(low_count = baseMean < low_count_threshold) %>%
  select(binder_id, condition, log2FC, baseMean, low_count) %>%
  arrange(condition, desc(log2FC))

hit_results
```

```{r final-plot}
hit_results %>%
  filter(condition == "toxin_plus_library_1mM") %>%
  slice_max(log2FC, n = 20) %>%
  ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = low_count)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), name = "low read count") +
  labs(x = "Binder", y = "log2FC vs. library-only (1 mM IPTG)",
       title = "Top 20 RcaT binders by protection") +
  theme_minimal()
```

```{r save-final-table}
write_csv(hit_results, "../results/final_binder_ranking.csv")
```

## 7. Comparison to the published plate-based screen (Sen2 retron)

Our collaborator ran an equivalent screen of these same binders with a
plate-based assay and shared the results (`data/plate_assay_reference_hits.tsv`),
filtered here to `retron == "Sen2"` (RcaT is the toxin from the Sen2
toxin-antitoxin system used in this project; the file's other retron, Eco9,
is a different system and isn't relevant here). Effect size > 1 means
blocking/protective, matching the sign of a positive `log2FC` in our own
result.

**This isn't an apples-to-apples comparison**: their effect sizes are
calculated against an IPTG-low baseline without further normalising
toxin+library against library-only the way we do here, and they additionally
correct both arms against a GFP control we don't have - so treat this as a
rough sanity check on direction and relative ranking, not a strict
validation of magnitude.

```{r load-plate-reference}
plate_reference <- read_tsv("../data/plate_assay_reference_hits.tsv", show_col_types = FALSE) %>%
  filter(retron == "Sen2") %>%
  transmute(genename, plate_pval = pval, plate_log2_efsize = log2(efsize), plate_hit = hit)

# Our binder_id values carry a naming-scheme prefix (`result_...` or
# `rcat-binder-..._result_<n>`) that the collaborator's gene names don't -
# strip it down to the same bare name before joining.
binder_name_map <- tibble(binder_id = unique(hit_results$binder_id)) %>%
  mutate(
    genename = binder_id %>%
      str_remove("^rcat-binder-") %>%
      str_remove("^result_") %>%
      str_replace("_result_", "_")
  )

comparison <- hit_results %>%
  left_join(binder_name_map, by = "binder_id") %>%
  left_join(plate_reference, by = "genename") %>%
  filter(!is.na(plate_log2_efsize))

comparison %>%
  ggplot(aes(x = plate_log2_efsize, y = log2FC, colour = plate_hit, shape = low_count)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey40", linetype = "dashed") +
  facet_wrap(~condition) +
  labs(x = "Published plate assay: log2(effect size), Sen2 (vs. IPTG-low)",
       y = "This screen: log2FC (toxin+library vs. library-only, matched IPTG)",
       title = "This screen vs. the published plate-based assay",
       colour = "Plate-assay hit call", shape = "Low read count") +
  theme_minimal()

# Does excluding the low-read-count binders (flagged in Section 6) change
# the picture? If most of the disagreement comes from binders Section 2
# already told us are read-starved, that's the library-complexity problem
# showing up here too, not a real disagreement between the two assays.
comparison %>%
  group_by(condition) %>%
  summarise(
    pearson_r_all             = cor(log2FC, plate_log2_efsize, use = "complete.obs"),
    pearson_r_excl_low_count  = cor(log2FC[!low_count], plate_log2_efsize[!low_count], use = "complete.obs"),
    n_binders_matched         = n(),
    n_excl_low_count          = sum(!low_count),
    .groups = "drop"
  )
```

```{r save-comparison-table}
write_csv(comparison, "../results/plate_assay_comparison.csv")
```

A positive correlation here is reassuring, but don't over-read a weak one:
this screen's baseline (library-only, matched IPTG) and the plate assay's
(IPTG-low, GFP-corrected) aren't quite the same comparison, both assays have
their own noise, and a binder missing from `plate_reference` above just
wasn't in that dataset (check `n_binders_matched` against the number of
viable binders from Section 2 to see how many actually got compared).

## Where to go from here

* The library-complexity check in Section 2 is worth flagging back to
  whoever built/pooled this plasmid library if it shows a handful of
  binders dominating the pool - it affects any future sequencing run of the
  same pool, not just this analysis, and a better-normalised pool (e.g.
  qPCR-checked equimolar pooling before transformation) would fix it at the
  source rather than needing to be corrected for downstream every time.
* Consider also looking at the 0.1 mM vs 1 mM IPTG comparison directly - a
  true binder should show a dose response (more enrichment at higher
  induction).
* No GFP (or other neutral overexpression) control is included in this
  sequencing-based run. Matching the IPTG level in Section 5 removes most of
  the overexpression-burden confound, but a real GFP control would remove it
  more rigorously and would bring this in line with the plate-based assay's
  full method.
* If a second replicate ever becomes available, re-running with both would
  let you add a genuine A-vs-B consistency check and a count-based test
  (DESeq2 or similar) that gives each binder a real uncertainty estimate
  instead of relying on the low-read-count flag alone.
