---
title: "RcaT binder pooled screen: DMS-style analysis"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does

We designed ~96 candidate binders against the RcaT toxin. Cells carrying the
pooled binder library were grown under 6 conditions (no inducer, toxin-only,
library-only at two IPTG levels, and toxin+library at two IPTG levels), each
in duplicate (A/B). RcaT kills the cell unless it's bound/inhibited, so **a
binder's relative read count after selection is a proxy for how well it
protects the cell** - good binders get relatively enriched when both the
toxin and the binder are induced together.

Steps below:

1. Load the counts table produced by `scripts/count_reads_per_binder.py`
   (`results/binder_counts.tsv`) and the sample metadata (`config/samples.tsv`).
2. QC: read depth per sample, how many binders are actually detected.
3. Normalise within each sample (sequencing depth differs sample to sample).
4. log2 fold-changes: first each induced condition vs. no-inducer baseline
   (sanity checks), then the condition that actually answers the biological
   question - toxin+library vs. toxin-only, i.e. does inducing the binder
   rescue growth under toxin challenge relative to toxin alone?
5. Check that replicates A and B agree.
6. Final ranked hit table.

```{r libraries}
library(tidyverse)   # readr, dplyr, tidyr, ggplot2, purrr
library(scales)      # nice axis labels
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

# scripts/count_reads_per_binder.py only writes a row when a binder has at
# least one read in a sample, so binder x sample combinations with zero
# reads are simply absent here. Fill those in explicitly as count = 0 -
# a binder legitimately dropping to zero in a condition (e.g. a non-binder
# under toxin+library selection) is exactly the kind of result we want to
# see below, not a row that silently disappears from the analysis.
counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, replicate, ara_pct, iptg_mM),
            by = "sample_id")

head(counts_long)
```

Reshape to a wide binder x sample matrix - this is the table students should
picture as "the" dataset from here on (rows = binders, columns = samples).

```{r wide-matrix}
counts_wide <- counts_long %>%
  select(binder_id, sample_id, count) %>%
  pivot_wider(names_from = sample_id, values_from = count, values_fill = 0)

counts_wide
```

## 2. QC: sequencing depth and binder detection

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition, replicate) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

If a sample has much lower depth than the others, everything downstream
(especially rare/low-abundance binders) will be noisier for that sample -
keep this table in mind when interpreting later plots. If you ran
`scripts/count_reads_per_binder.py`, also check the companion
`results/binder_counts_qc.tsv` for how many read pairs were dropped as
unmapped / low-MAPQ / mate-mismatch per sample - a sample with a lot of
`mate_reference_mismatch` hits suggests some binders may share long
identical stretches of sequence and can't always be told apart from a single
read pair.

## 3. Normalisation

Sequencing depth differs between samples, so raw counts aren't directly
comparable. We convert each binder's count to its **relative frequency
within the sample** (equivalent to counts-per-million, just expressed as a
fraction) - this is the natural normalisation for a competitive pooled assay,
since what we actually care about is what fraction of the surviving
population each binder makes up, not its absolute read count.

A small pseudocount is added before taking logs so that binders with zero
reads in a given sample don't produce `-Inf`.

```{r normalise}
pseudocount <- 1

counts_norm <- counts_long %>%
  group_by(sample_id) %>%
  mutate(
    total_reads = sum(count),
    freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_long$binder_id))
  ) %>%
  ungroup()

counts_norm %>% select(sample_id, binder_id, count, total_reads, freq) %>% head()
```

## 4. log2 fold-changes

We average the two replicates' normalised frequency for each condition
first, then take log2 ratios between conditions. (Section 5 checks the
replicates actually agree with each other before we lean on this average.)

```{r condition-means}
condition_means <- counts_norm %>%
  group_by(condition, binder_id) %>%
  summarise(mean_freq = mean(freq), .groups = "drop")

condition_means %>% head()
```

### 4a. Sanity-check comparisons: each induced condition vs. no-inducer

These aren't the final answer, but they're a useful check that induction is
doing *something* (e.g. toxin-only should broadly suppress growth relative
to no-inducer, roughly uniformly across binders since none are expressed yet
- IPTG is 0 in that condition).

```{r lfc-vs-baseline}
baseline <- condition_means %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, baseline_freq = mean_freq)

lfc_vs_baseline <- condition_means %>%
  filter(condition != "no_inducer") %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / baseline_freq))

lfc_vs_baseline %>%
  ggplot(aes(x = binder_id, y = log2FC)) +
  geom_col() +
  facet_wrap(~condition, ncol = 1) +
  labs(x = "Binder", y = "log2FC vs. no_inducer", title = "Each condition vs. no-inducer baseline") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

### 4b. The actual question: toxin+library vs. toxin-only

This isolates the effect of *inducing the binder* while the toxin is
present, which is what should separate good binders (protective, so
enriched) from the rest. We do this at both IPTG concentrations.

```{r lfc-key-comparison}
toxin_only_mean <- condition_means %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, toxin_only_freq = mean_freq)

key_comparison <- condition_means %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_mean, by = "binder_id") %>%
  mutate(log2FC_vs_toxinOnly = log2(mean_freq / toxin_only_freq))

key_comparison %>%
  ggplot(aes(x = reorder(binder_id, log2FC_vs_toxinOnly), y = log2FC_vs_toxinOnly, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs. toxin-only",
       title = "Enrichment of each binder when co-induced with toxin",
       subtitle = "Positive = binder rescues growth (protective) relative to toxin alone") +
  theme_minimal()
```

## 5. Replicate consistency

Before trusting the log2FC table above, check that replicate A and
replicate B agree - both on raw-ish normalised frequency (library
complexity/correlation) and on the derived log2FC (biological
reproducibility of the actual result we care about).

```{r replicate-freq-correlation}
rep_wide <- counts_norm %>%
  select(sample_id, binder_id, freq, condition, replicate) %>%
  pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = freq)

rep_wide %>%
  ggplot(aes(x = A, y = B)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  scale_x_log10(labels = label_number()) +
  scale_y_log10(labels = label_number()) +
  facet_wrap(~condition) +
  labs(title = "Replicate A vs B: normalised binder frequency", x = "Rep A freq", y = "Rep B freq") +
  theme_minimal()

rep_wide %>%
  group_by(condition) %>%
  summarise(pearson_r = cor(log10(A), log10(B), use = "complete.obs"), .groups = "drop")
```

Now the same check on the log2FC that actually feeds the hit-calling below -
compute log2FC(toxin+library vs toxin-only) **per replicate** (not averaged)
and compare A vs B directly.

```{r replicate-lfc-consistency}
per_rep_freq <- counts_norm %>%
  select(sample_id, binder_id, condition, replicate, freq)

toxin_only_per_rep <- per_rep_freq %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, replicate, toxin_only_freq = freq)

lfc_per_rep <- per_rep_freq %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_per_rep, by = c("binder_id", "replicate")) %>%
  mutate(log2FC = log2(freq / toxin_only_freq))

lfc_per_rep_wide <- lfc_per_rep %>%
  pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = log2FC)

lfc_per_rep_wide %>%
  ggplot(aes(x = A, y = B)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~condition) +
  labs(title = "Replicate A vs B: log2FC (toxin+library vs toxin-only)",
       x = "Rep A log2FC", y = "Rep B log2FC") +
  theme_minimal()

lfc_per_rep_wide %>%
  group_by(condition) %>%
  summarise(pearson_r = cor(A, B, use = "complete.obs"), .groups = "drop")
```

**A binder is only worth trusting as a hit if its two replicates agree** -
flag anything where A and B disagree in sign or differ by a lot, and treat
those as inconclusive rather than as hits or non-hits.

```{r flag-inconsistent}
consistency_threshold <- 1  # max allowed |A - B| in log2FC units; tune by eye from the plot above

lfc_flagged <- lfc_per_rep_wide %>%
  mutate(
    same_direction = sign(A) == sign(B),
    abs_diff       = abs(A - B),
    consistent     = same_direction & abs_diff <= consistency_threshold
  )

lfc_flagged %>% count(condition, consistent)
```

## 6. Final ranked hit table

Average the two replicates' log2FC (toxin+library vs toxin-only) for
consistent binders only, and rank.

```{r final-hits}
final_hits <- lfc_flagged %>%
  filter(consistent) %>%
  mutate(mean_log2FC = (A + B) / 2) %>%
  select(binder_id, condition, mean_log2FC, A, B) %>%
  arrange(condition, desc(mean_log2FC))

final_hits
```

```{r final-plot}
final_hits %>%
  filter(condition == "toxin_plus_library_1mM") %>%
  slice_max(mean_log2FC, n = 20) %>%
  ggplot(aes(x = reorder(binder_id, mean_log2FC), y = mean_log2FC)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmin(A, B), ymax = pmax(A, B)), width = 0.2) +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs toxin-only (1 mM IPTG)",
       title = "Top 20 RcaT binders by replicate-consistent enrichment") +
  theme_minimal()
```

```{r save-final-table}
write_csv(final_hits, "../results/final_binder_ranking.csv")
```

## Where to go from here

* This uses a simple frequency-ratio log2FC, which is easy to reason about
  but doesn't model count-based variance the way tools like DESeq2/edgeR do
  (which down-weight noisy, low-count binders and give you a p-value per
  binder). If you want to go further, feeding `counts_wide` into DESeq2 with
  a design of `~ condition` and pulling the `toxin_plus_library_*` vs.
  `toxin_only` contrast is a natural next step.
* Consider also looking at the 0.1 mM vs 1 mM IPTG comparison directly - a
  true binder should show a dose response (more enrichment at higher
  induction), which is another consistency check independent of replicates.
