---
title: "RcaT binder pooled screen: DMS-style analysis"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does

We designed 88 candidate binders against the RcaT toxin. Cells carrying the
pooled binder library were grown under 6 conditions (no inducer, toxin-only,
library-only at two IPTG levels, and toxin+library at two IPTG levels), each
in duplicate (A/B). RcaT kills the cell unless it's bound/inhibited, so **a
binder's relative read count after selection is a proxy for how well it
protects the cell** - good binders get relatively enriched when both the
toxin and the binder are induced together.

This mirrors the lab's existing plate-based version of the same assay, and
uses the same three roles for the 6 conditions:

- **no_inducer** (antibiotic only): a viability check, not a comparison
  baseline. A binder that doesn't survive plain antibiotic selection isn't a
  usable data point for anything downstream, so it's dropped here rather
  than showing up later as a meaningless fold-change.
- **library_only** (IPTG only, two levels, no toxin): tells you the *innate*
  toxicity of overexpressing each binder on its own, with no RcaT present -
  reported for context, not used to correct the main result.
- **toxin_only** vs. **toxin_plus_library**: the actual comparison. Does
  inducing the binder rescue growth under toxin challenge, relative to toxin
  alone? The plate-based version of this assay additionally normalises both
  arms against a GFP-only control plasmid; we don't have an equivalent GFP
  control in this sequencing-based setup, so this is the direct,
  un-normalised version of that same ratio (see the caveat in Section 5).

Steps below:

1. Load the counts table produced by `scripts/count_reads_per_binder.py`
   (`results/binder_counts.tsv`) and the sample metadata (`config/samples.tsv`).
2. QC: read depth per sample, how many binders are detected, and a viability
   filter that drops binders that don't survive no-inducer selection.
3. Normalise within each sample (sequencing depth differs sample to sample).
4. Innate toxicity: library-only vs. no-inducer, as context only.
5. The actual question: toxin+library vs. toxin-only.
6. Check that replicates A and B agree.
7. Final ranked hit table.
8. Compare against the collaborator's published plate-based screen of the
   same binders (Sen2 retron), as an independent sanity check.

```{r libraries}
library(tidyverse)   # readr, dplyr, tidyr, ggplot2, purrr
library(scales)      # nice axis labels
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

# scripts/count_reads_per_binder.py only writes a row when a binder has at
# least one read in a sample, so binder x sample combinations with zero
# reads are simply absent here. Fill those in explicitly as count = 0 -
# a binder legitimately dropping to zero in a condition (e.g. a non-binder
# under toxin+library selection) is exactly the kind of result we want to
# see below, not a row that silently disappears from the analysis.
counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, replicate, ara_pct, iptg_mM),
            by = "sample_id")

# Section 5 onward compares replicate A against replicate B directly - only
# possible if both actually exist in this run (e.g. not yet true for a
# single-replicate test run before the full experiment). Everything before
# Section 5 works identically either way, since averaging one value is a
# no-op.
has_replicates <- all(c("A", "B") %in% unique(samples$replicate))
if (!has_replicates) {
  message("Only replicate(s) ", paste(unique(samples$replicate), collapse = ", "),
          " present - Section 5's A-vs-B consistency checks will be skipped, and the ",
          "final hit table will be marked provisional (not replicate-validated). ",
          "Re-run this notebook once both replicates exist.")
}

head(counts_long)
```

Reshape to a wide binder x sample matrix - this is the table students should
picture as "the" dataset from here on (rows = binders, columns = samples).

```{r wide-matrix}
counts_wide <- counts_long %>%
  select(binder_id, sample_id, count) %>%
  pivot_wider(names_from = sample_id, values_from = count, values_fill = 0)

counts_wide
```

## 2. QC: sequencing depth and binder detection

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition, replicate) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

If a sample has much lower depth than the others, everything downstream
(especially rare/low-abundance binders) will be noisier for that sample -
keep this table in mind when interpreting later plots. If you ran
`scripts/count_reads_per_binder.py`, also check the companion
`results/binder_counts_qc.tsv` for how many read pairs were dropped as
unmapped / low-MAPQ / mate-mismatch per sample - a sample with a lot of
`mate_reference_mismatch` hits suggests some binders may share long
identical stretches of sequence and can't always be told apart from a single
read pair.

### Viability filter: drop binders that don't survive no-inducer selection

The plate-based version of this assay uses the no-inducer condition
(antibiotic only, no Ara/IPTG) purely as a check of which plasmids survive at
all - a binder that's already essentially gone before any toxin or induction
step isn't a meaningful data point for anything downstream, and would just
show up later as a huge, meaningless fold-change from a near-empty
denominator. We do the same thing here: drop any binder whose average read
count under `no_inducer` falls below a minimum threshold, before any
normalisation or comparison.

```{r viability-filter}
min_no_inducer_reads <- 10  # tune by eye against the histogram below

no_inducer_counts <- counts_long %>%
  filter(condition == "no_inducer") %>%
  group_by(binder_id) %>%
  summarise(mean_no_inducer_count = mean(count), .groups = "drop")

no_inducer_counts %>%
  ggplot(aes(x = mean_no_inducer_count)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = min_no_inducer_reads, linetype = "dashed", colour = "red") +
  labs(x = "Mean no_inducer read count", y = "Number of binders",
       title = "Choosing the viability threshold",
       subtitle = paste("Dashed line: current threshold =", min_no_inducer_reads)) +
  theme_minimal()

dead_binders <- no_inducer_counts %>%
  filter(mean_no_inducer_count < min_no_inducer_reads) %>%
  pull(binder_id)

if (length(dead_binders) > 0) {
  message(length(dead_binders), " binder(s) dropped as non-viable (mean no_inducer count < ",
          min_no_inducer_reads, "): ", paste(dead_binders, collapse = ", "))
} else {
  message("No binders dropped - all binders survive no-inducer selection above threshold.")
}

counts_long <- counts_long %>% filter(!(binder_id %in% dead_binders))
```

Every plot and table from here on only includes binders that actually grow
under plain antibiotic selection.

## 3. Normalisation

Sequencing depth differs between samples, so raw counts aren't directly
comparable. We convert each binder's count to its **relative frequency
within the sample** (equivalent to counts-per-million, just expressed as a
fraction) - this is the natural normalisation for a competitive pooled assay,
since what we actually care about is what fraction of the surviving
population each binder makes up, not its absolute read count.

A small pseudocount is added before taking logs so that binders with zero
reads in a given sample don't produce `-Inf`.

```{r normalise}
pseudocount <- 1

counts_norm <- counts_long %>%
  group_by(sample_id) %>%
  mutate(
    total_reads = sum(count),
    freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_long$binder_id))
  ) %>%
  ungroup()

counts_norm %>% select(sample_id, binder_id, count, total_reads, freq) %>% head()
```

## 4. Innate toxicity of overexpression (reference only)

We average the two replicates' normalised frequency for each condition
first, then take log2 ratios between conditions. (Section 6 checks the
replicates actually agree with each other before we lean on this average.)

```{r condition-means}
condition_means <- counts_norm %>%
  group_by(condition, binder_id) %>%
  summarise(mean_freq = mean(freq), .groups = "drop")

condition_means %>% head()
```

This isn't part of the actual hit-calling comparison in Section 5 - it's
here purely as context, mirroring what the plate-based assay calls "innate
toxicity of the plasmids": how much does overexpressing each binder cost the
cell on its own, with no toxin present at all (`library_only` vs.
`no_inducer`, at both IPTG levels)?

```{r innate-toxicity}
baseline <- condition_means %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, baseline_freq = mean_freq)

innate_toxicity <- condition_means %>%
  filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
  left_join(baseline, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / baseline_freq))

innate_toxicity %>%
  ggplot(aes(x = binder_id, y = log2FC)) +
  geom_col() +
  facet_wrap(~condition, ncol = 1) +
  labs(x = "Binder", y = "log2FC vs. no_inducer",
       title = "Innate toxicity of overexpression (no toxin present)",
       subtitle = "Reference only - not used in the hit-calling comparison below") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

Binders far to the negative side here are costly to overexpress regardless
of RcaT - worth keeping in mind if they also show up near the top of the hit
table below (see the caveat in Section 5).

## 5. The actual question: toxin+library vs. toxin-only

This is the direct comparison the plate-based version of this assay makes
(there, both arms are additionally normalised against a GFP-only control
plasmid to correct for general overexpression burden; we don't have an
equivalent GFP control in this sequencing-based setup, so this is the
un-normalised version of that same ratio). We do this at both IPTG
concentrations.

```{r key-comparison}
toxin_only_mean <- condition_means %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, toxin_only_freq = mean_freq)

key_comparison <- condition_means %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_mean, by = "binder_id") %>%
  mutate(log2FC = log2(mean_freq / toxin_only_freq))

key_comparison %>%
  ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = condition)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = "Binder", y = "log2FC vs. toxin-only",
       title = "Enrichment of each binder when co-induced with toxin",
       subtitle = "Positive = binder rescues growth (protective) relative to toxin alone") +
  theme_minimal()
```

**Caveat:** without a GFP (or other neutral-overexpression) control, this
comparison can't fully separate RcaT-specific protection from a binder
simply being cheap to overexpress. Cross-check any top hit against the
innate toxicity plot above (a hit that's also very costly to overexpress on
its own is less convincing) and against the 0.1 mM vs. 1 mM dose response
below (a true binder should show *more* protection at higher induction; a
burden artifact generally doesn't).

## 6. Replicate consistency

Before trusting the log2FC table above, check that replicate A and
replicate B agree - both on raw-ish normalised frequency (library
complexity/correlation) and on the derived log2FC (biological
reproducibility of the actual result we care about).

```{r replicate-freq-correlation}
if (has_replicates) {
  rep_wide <- counts_norm %>%
    select(sample_id, binder_id, freq, condition, replicate) %>%
    pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = freq)

  print(
    rep_wide %>%
      ggplot(aes(x = A, y = B)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      scale_x_log10(labels = label_number()) +
      scale_y_log10(labels = label_number()) +
      facet_wrap(~condition) +
      labs(title = "Replicate A vs B: normalised binder frequency", x = "Rep A freq", y = "Rep B freq") +
      theme_minimal()
  )

  rep_wide %>%
    group_by(condition) %>%
    summarise(pearson_r = cor(log10(A), log10(B), use = "complete.obs"), .groups = "drop")
} else {
  message("Skipping - only one replicate present.")
}
```

Now the same check on the actual `log2FC` from Section 5 - compute it **per
replicate** (not averaged) and compare A vs B directly.

```{r replicate-lfc-consistency}
per_rep_freq <- counts_norm %>%
  select(sample_id, binder_id, condition, replicate, freq)

toxin_only_per_rep <- per_rep_freq %>%
  filter(condition == "toxin_only") %>%
  select(binder_id, replicate, toxin_only_freq = freq)

lfc_per_rep <- per_rep_freq %>%
  filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
  left_join(toxin_only_per_rep, by = c("binder_id", "replicate")) %>%
  mutate(log2FC = log2(freq / toxin_only_freq))

if (has_replicates) {
  lfc_per_rep_wide <- lfc_per_rep %>%
    pivot_wider(id_cols = c(binder_id, condition), names_from = replicate, values_from = log2FC)

  print(
    lfc_per_rep_wide %>%
      ggplot(aes(x = A, y = B)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
      facet_wrap(~condition) +
      labs(title = "Replicate A vs B: log2FC (toxin+library vs toxin-only)",
           x = "Rep A log2FC", y = "Rep B log2FC") +
      theme_minimal()
  )

  lfc_per_rep_wide %>%
    group_by(condition) %>%
    summarise(pearson_r = cor(A, B, use = "complete.obs"), .groups = "drop")
} else {
  message("Skipping - only one replicate present.")
}
```

**A binder is only worth trusting as a hit if its two replicates agree** -
flag anything where A and B disagree in sign or differ by a lot, and treat
those as inconclusive rather than as hits or non-hits. (Skipped entirely if
only one replicate exists - there's nothing to check agreement against yet.)

```{r flag-inconsistent}
consistency_threshold <- 1  # max allowed |A - B| in log2FC units; tune by eye from the plot above

if (has_replicates) {
  lfc_flagged <- lfc_per_rep_wide %>%
    mutate(
      same_direction = sign(A) == sign(B),
      abs_diff       = abs(A - B),
      consistent     = same_direction & abs_diff <= consistency_threshold
    )

  lfc_flagged %>% count(condition, consistent)
} else {
  message("Skipping - only one replicate present.")
}
```

## 7. Final ranked hit table

With both replicates, average the log2FC (toxin+library vs. toxin-only) for
consistent binders only, and rank. With a single replicate, rank directly on
that replicate's value instead - but mark the table **provisional**: nothing
here has been checked for reproducibility yet, so treat it as "worth a
closer look", not "confirmed hits", until replicate B exists.

```{r final-hits}
if (has_replicates) {
  final_hits <- lfc_flagged %>%
    filter(consistent) %>%
    mutate(mean_log2FC = (A + B) / 2, provisional = FALSE) %>%
    select(binder_id, condition, mean_log2FC, A, B, provisional) %>%
    arrange(condition, desc(mean_log2FC))
} else {
  final_hits <- lfc_per_rep %>%
    mutate(mean_log2FC = log2FC, A = log2FC, B = NA_real_, provisional = TRUE) %>%
    select(binder_id, condition, mean_log2FC, A, B, provisional) %>%
    arrange(condition, desc(mean_log2FC))
}

final_hits
```

```{r final-plot}
final_hits %>%
  filter(condition == "toxin_plus_library_1mM") %>%
  slice_max(mean_log2FC, n = 20) %>%
  ggplot(aes(x = reorder(binder_id, mean_log2FC), y = mean_log2FC, fill = provisional)) +
  geom_col() +
  { if (has_replicates) geom_errorbar(aes(ymin = pmin(A, B), ymax = pmax(A, B)), width = 0.2) } +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), guide = if (has_replicates) "none" else "legend") +
  labs(x = "Binder", y = "log2FC vs. toxin-only (1 mM IPTG)",
       title = paste("Top 20 RcaT binders by", if (has_replicates) "replicate-consistent protection" else "protection (provisional - single replicate)")) +
  theme_minimal()
```

```{r save-final-table}
write_csv(final_hits, "../results/final_binder_ranking.csv")
```

## 8. Comparison to the published plate-based screen (Sen2 retron)

Our collaborator ran an equivalent screen of these same binders with a
plate-based assay and shared the results (`data/plate_assay_reference_hits.tsv`),
filtered here to `retron == "Sen2"` (RcaT is the toxin from the Sen2
toxin-antitoxin system used in this project; the file's other retron, Eco9,
is a different system and isn't relevant here). Effect size > 1 means
blocking/protective, matching the sign of a positive `log2FC` in our own
result.

**This isn't an apples-to-apples comparison**: their effect sizes are
calculated against an IPTG-low baseline, not our toxin-only baseline, so
treat this as a rough sanity check on direction and relative ranking, not a
strict validation of magnitude.

```{r load-plate-reference}
plate_reference <- read_tsv("../data/plate_assay_reference_hits.tsv", show_col_types = FALSE) %>%
  filter(retron == "Sen2") %>%
  transmute(genename, plate_pval = pval, plate_log2_efsize = log2(efsize), plate_hit = hit)

# Our binder_id values carry a naming-scheme prefix (`result_...` or
# `rcat-binder-..._result_<n>`) that the collaborator's gene names don't -
# strip it down to the same bare name before joining.
binder_name_map <- tibble(binder_id = unique(key_comparison$binder_id)) %>%
  mutate(
    genename = binder_id %>%
      str_remove("^rcat-binder-") %>%
      str_remove("^result_") %>%
      str_replace("_result_", "_")
  )

comparison <- key_comparison %>%
  left_join(binder_name_map, by = "binder_id") %>%
  left_join(plate_reference, by = "genename") %>%
  filter(!is.na(plate_log2_efsize))

comparison %>%
  ggplot(aes(x = plate_log2_efsize, y = log2FC, colour = plate_hit)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey40", linetype = "dashed") +
  facet_wrap(~condition) +
  labs(x = "Published plate assay: log2(effect size), Sen2 (vs. IPTG-low)",
       y = "This screen: log2FC (toxin+library vs. toxin-only)",
       title = "This screen vs. the published plate-based assay",
       colour = "Plate-assay hit call") +
  theme_minimal()

comparison %>%
  group_by(condition) %>%
  summarise(pearson_r = cor(log2FC, plate_log2_efsize, use = "complete.obs"),
            n_binders_matched = n(), .groups = "drop")
```

```{r save-comparison-table}
write_csv(comparison, "../results/plate_assay_comparison.csv")
```

A positive correlation here is reassuring, but don't over-read a weak one:
this screen's baseline (toxin-only) and the plate assay's (IPTG-low) aren't
the same condition, both assays have their own noise, and a binder missing
from `plate_reference` above just wasn't in that dataset (check
`n_binders_matched` against the number of viable binders from Section 2 to
see how many actually got compared).

## Where to go from here

* This uses a simple frequency-ratio log2FC, which is easy to reason about
  but doesn't model count-based variance the way tools like DESeq2/edgeR do
  (which down-weight noisy, low-count binders and give you a p-value per
  binder). If you want to go further, feeding `counts_wide` into DESeq2 with
  a design of `~ condition` and pulling the `toxin_plus_library_*` vs.
  `toxin_only` contrast is a natural next step.
* Consider also looking at the 0.1 mM vs 1 mM IPTG comparison directly - a
  true binder should show a dose response (more enrichment at higher
  induction), which is another consistency check independent of replicates.
* No GFP (or other neutral overexpression) control is included in this
  sequencing-based run. If one gets added to a future run, normalising both
  `toxin_only` and `toxin_plus_library` against it before taking the log2FC
  in Section 5 would bring this in line with the plate-based assay's full
  method and remove the caveat above.
