---
title: "RcaT binder pooled screen: DMS-style analysis (high/low-expression batches)"
author: "PhD Primer - Dorrity/Mahamid/Eustermann groups"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
```

## 0. What this notebook does differently

This is a variant of `DMS_analysis.Rmd` (see that notebook for the full
walkthrough of the experiment, the conditions, and the basic logic - it's
not repeated here). The difference is in how it handles something that
notebook's library-complexity check (Section 2) turned up: in this pool, a
handful of binders dominate the read counts in every sample, while most of
the other ~80 are read-starved by comparison. Because every binder's count
is really a *share* of a fixed sequencing total, those dominant binders
distort the normalisation for everyone else - even a binder whose own
abundance didn't really change can look enriched or depleted just because
the dominant clones' share shifted around it.

**The fix here**: split the panel into a high-expressing batch and a
low-expressing batch *before* any normalisation or hit-calling, and run the
entire comparison separately within each batch. Low-expressing binders are
never thrown out - they're just compared only against other binders at a
similar abundance, instead of being compositionally squeezed by a few
dominant clones that have nothing to do with them.

**What this does and doesn't fix**: splitting removes the dominant clones
from the low-expressing batch's own normalisation, which is a real
distortion. It does *not* manufacture extra sequencing depth - a binder with
genuinely few reads still has genuinely few reads, and is still flagged
`low_count` for it. Splitting fixes a normalisation bias, not a
sequencing-depth limit.

**No replicates**: like `DMS_analysis.Rmd`, this assumes a single sample per
condition, so every value below is a one-off point estimate - there's no
A-vs-B consistency check, and no count-based statistical test (DESeq2 and
similar tools need at least 2 replicates per condition to estimate
dispersion, so none is used here).

```{r libraries}
library(tidyverse)
library(scales)
```

## 1. Load data

```{r load-data}
counts_long <- read_tsv("../results/binder_counts.tsv", show_col_types = FALSE)
samples     <- read_tsv("../config/samples.tsv", show_col_types = FALSE)

counts_long <- counts_long %>%
  complete(binder_id, sample_id = samples$sample_id, fill = list(count = 0)) %>%
  left_join(samples %>% select(sample_id, condition, ara_pct, iptg_mM),
            by = "sample_id")

head(counts_long)
```

## 2. QC: sequencing depth, library complexity, viability

Same QC as `DMS_analysis.Rmd` - included here so the batch split below has
a clean, already-filtered starting point.

```{r qc-depth}
depth <- counts_long %>%
  group_by(sample_id, condition) %>%
  summarise(total_reads = sum(count), n_binders_detected = sum(count > 0), .groups = "drop")

depth %>%
  ggplot(aes(x = reorder(sample_id, total_reads), y = total_reads, fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total usable read pairs", title = "Sequencing depth per sample") +
  theme_minimal()
```

```{r qc-table}
depth
```

### Library complexity: how evenly is the pool represented?

```{r library-complexity}
complexity <- counts_long %>%
  group_by(sample_id, condition) %>%
  mutate(sample_total = sum(count)) %>%
  arrange(sample_id, desc(count)) %>%
  mutate(rank = row_number(), cum_share = cumsum(count) / sample_total) %>%
  ungroup()

complexity %>%
  ggplot(aes(x = rank, y = cum_share, group = sample_id, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  labs(x = "Binders, ranked by abundance (most abundant first)",
       y = "Cumulative share of reads",
       title = "Library complexity per sample",
       subtitle = "A steep early climb to the dashed line (90%) means a handful of binders dominate the pool") +
  theme_minimal()
```

This is exactly the pattern that motivates the batch split below - if it's
flat/gradual here (no dominant clones), splitting into batches won't change
much and you may as well use the unsplit `DMS_analysis.Rmd` instead.

### Viability filter: drop binders that don't survive no-inducer selection

```{r viability-filter}
min_no_inducer_reads <- 10  # tune by eye against the histogram below

no_inducer_counts <- counts_long %>%
  filter(condition == "no_inducer") %>%
  select(binder_id, no_inducer_count = count)

no_inducer_counts %>%
  ggplot(aes(x = no_inducer_count)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = min_no_inducer_reads, linetype = "dashed", colour = "red") +
  labs(x = "no_inducer read count", y = "Number of binders",
       title = "Choosing the viability threshold",
       subtitle = paste("Dashed line: current threshold =", min_no_inducer_reads)) +
  theme_minimal()

dead_binders <- no_inducer_counts %>%
  filter(no_inducer_count < min_no_inducer_reads) %>%
  pull(binder_id)

if (length(dead_binders) > 0) {
  message(length(dead_binders), " binder(s) dropped as non-viable (no_inducer count < ",
          min_no_inducer_reads, "): ", paste(dead_binders, collapse = ", "))
} else {
  message("No binders dropped - all binders survive no-inducer selection above threshold.")
}

counts_long <- counts_long %>% filter(!(binder_id %in% dead_binders))
```

### Splitting into high- and low-expressing batches

Rank binders by their overall abundance (mean raw read count across *all*
samples and conditions - not just one condition, since Section 2's library
complexity check showed the same handful of binders dominate every
condition about equally). Binders at or above the threshold go in the
high-expressing batch; everyone else goes in the low-expressing batch.
Nothing is excluded here - every viable binder lands in exactly one batch.

```{r expression-batch-split}
high_expr_threshold <- 1000  # mean read count across all samples; tune by eye against the histogram below

overall_abundance <- counts_long %>%
  group_by(binder_id) %>%
  summarise(mean_count_overall = mean(count), .groups = "drop")

overall_abundance %>%
  ggplot(aes(x = mean_count_overall)) +
  geom_histogram(bins = 30) +
  scale_x_log10(labels = label_number()) +
  geom_vline(xintercept = high_expr_threshold, linetype = "dashed", colour = "red") +
  labs(x = "Mean read count across all samples (log scale)", y = "Number of binders",
       title = "Choosing the high/low expression split",
       subtitle = paste("Dashed line: current threshold =", high_expr_threshold)) +
  theme_minimal()

expr_batch <- overall_abundance %>%
  mutate(batch = if_else(mean_count_overall >= high_expr_threshold, "high", "low"))

high_expr_binders <- expr_batch %>% filter(batch == "high") %>% pull(binder_id)
low_expr_binders  <- expr_batch %>% filter(batch == "low") %>% pull(binder_id)

message(length(high_expr_binders), " binder(s) assigned to the high-expressing batch, ",
        length(low_expr_binders), " to the low-expressing batch.")

# Sanity check: the high batch should hold most of the READS; the low batch
# should hold most of the BINDERS. That mismatch is the whole reason to split.
counts_long %>%
  left_join(expr_batch %>% select(binder_id, batch), by = "binder_id") %>%
  group_by(batch) %>%
  summarise(n_binders = n_distinct(binder_id), total_reads = sum(count), .groups = "drop") %>%
  mutate(share_of_reads = total_reads / sum(total_reads))
```

## 3. The per-batch pipeline

Everything from here on (normalisation, innate toxicity, the key comparison,
the final hit table) is identical in logic to `DMS_analysis.Rmd` Sections
3-6 - it's written once as a function and run twice (Sections 4 and 5), so
the two batches get exactly the same treatment and nothing diverges by
accident. The key point: every `group_by(sample_id)` normalisation step only
ever sees the binders in *that* batch - a low-expressing binder's frequency
is computed against the low batch's own total, never diluted by the high
batch's reads.

```{r batch-pipeline-functions}
pseudocount <- 1
low_count_threshold <- 20  # flag (don't exclude) binders averaging fewer reads than this - tune by eye

run_batch_pipeline <- function(counts_batch) {
  counts_norm <- counts_batch %>%
    group_by(sample_id) %>%
    mutate(
      total_reads = sum(count),
      freq        = (count + pseudocount) / (total_reads + pseudocount * n_distinct(counts_batch$binder_id))
    ) %>%
    ungroup()

  condition_freq <- counts_norm %>%
    select(condition, binder_id, freq)

  baseline <- condition_freq %>%
    filter(condition == "no_inducer") %>%
    select(binder_id, baseline_freq = freq)

  innate_toxicity <- condition_freq %>%
    filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
    left_join(baseline, by = "binder_id") %>%
    mutate(log2FC = log2(freq / baseline_freq))

  p_innate_toxicity <- innate_toxicity %>%
    ggplot(aes(x = binder_id, y = log2FC)) +
    geom_col() +
    facet_wrap(~condition, ncol = 1) +
    labs(x = "Binder", y = "log2FC vs. no_inducer",
         title = "Innate toxicity of overexpression (no toxin present)",
         subtitle = "Reference only - not used in hit-calling") +
    theme_minimal() +
    theme(axis.text.x = element_blank())

  # The actual comparison: toxin+library vs. library-only, matched IPTG
  # level. Both arms have the binder induced at the same level, so each
  # binder's own overexpression cost is present on both sides of the ratio
  # and should mostly cancel out - what's left is mainly the effect of
  # adding the toxin on top of an already-expressed binder.
  library_only_freq <- condition_freq %>%
    filter(condition %in% c("library_only_0.1mM", "library_only_1mM")) %>%
    mutate(iptg_level = str_remove(condition, "library_only_")) %>%
    select(binder_id, iptg_level, library_only_freq = freq)

  key_comparison <- condition_freq %>%
    filter(condition %in% c("toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
    mutate(iptg_level = str_remove(condition, "toxin_plus_library_")) %>%
    left_join(library_only_freq, by = c("binder_id", "iptg_level")) %>%
    mutate(log2FC = log2(freq / library_only_freq)) %>%
    select(-iptg_level)

  p_key_comparison <- key_comparison %>%
    ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = condition)) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(x = "Binder", y = "log2FC vs. library-only (same IPTG level)",
         title = "Enrichment when co-induced with toxin (this batch only)",
         subtitle = "Positive = binder does better under toxin challenge than its own baseline without toxin") +
    theme_minimal()

  # Final hit table for this batch - flag (don't exclude) low-read-depth binders
  raw_mean_count <- counts_batch %>%
    filter(condition %in% c("library_only_0.1mM", "library_only_1mM",
                             "toxin_plus_library_0.1mM", "toxin_plus_library_1mM")) %>%
    group_by(binder_id) %>%
    summarise(baseMean = mean(count), .groups = "drop")

  hit_results <- key_comparison %>%
    left_join(raw_mean_count, by = "binder_id") %>%
    mutate(low_count = baseMean < low_count_threshold) %>%
    select(binder_id, condition, log2FC, baseMean, low_count) %>%
    arrange(condition, desc(log2FC))

  p_final <- hit_results %>%
    filter(condition == "toxin_plus_library_1mM") %>%
    slice_max(log2FC, n = 20) %>%
    ggplot(aes(x = reorder(binder_id, log2FC), y = log2FC, fill = low_count)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "grey60"), name = "low read count") +
    labs(x = "Binder", y = "log2FC vs. library-only (1 mM IPTG)",
         title = "Top binders by protection (this batch only)") +
    theme_minimal()

  list(
    innate_toxicity = innate_toxicity, p_innate_toxicity = p_innate_toxicity,
    key_comparison = key_comparison, p_key_comparison = p_key_comparison,
    hit_results = hit_results, p_final = p_final
  )
}

compare_to_plate <- function(hit_results, plate_reference, batch_label) {
  binder_name_map <- tibble(binder_id = unique(hit_results$binder_id)) %>%
    mutate(
      genename = binder_id %>%
        str_remove("^rcat-binder-") %>%
        str_remove("^result_") %>%
        str_replace("_result_", "_")
    )

  comparison <- hit_results %>%
    left_join(binder_name_map, by = "binder_id") %>%
    left_join(plate_reference, by = "genename") %>%
    filter(!is.na(plate_log2_efsize))

  p <- comparison %>%
    ggplot(aes(x = plate_log2_efsize, y = log2FC, colour = plate_hit, shape = low_count)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linetype = "dashed") +
    facet_wrap(~condition) +
    labs(x = "Published plate assay: log2(effect size), Sen2 (vs. IPTG-low)",
         y = paste0(batch_label, "-batch log2FC (toxin+library vs. library-only, matched IPTG)"),
         title = paste0("This screen (", batch_label, " batch) vs. the published plate-based assay"),
         colour = "Plate-assay hit call", shape = "Low read count") +
    theme_minimal()

  cor_summary <- comparison %>%
    group_by(condition) %>%
    summarise(
      pearson_r_all            = cor(log2FC, plate_log2_efsize, use = "complete.obs"),
      pearson_r_excl_low_count = cor(log2FC[!low_count], plate_log2_efsize[!low_count], use = "complete.obs"),
      n_binders_matched        = n(),
      n_excl_low_count         = sum(!low_count),
      .groups = "drop"
    ) %>%
    mutate(batch = batch_label, .before = 1)

  list(comparison = comparison, plot = p, cor_summary = cor_summary)
}
```

## 4. High-expressing batch

```{r high-batch-run}
high_batch_counts <- counts_long %>% filter(binder_id %in% high_expr_binders)
high <- run_batch_pipeline(high_batch_counts)
```

```{r high-innate-toxicity}
high$p_innate_toxicity
```

```{r high-key-comparison}
high$p_key_comparison
```

**Caution**: this batch is, by construction, a small handful of binders -
any correlation computed from it below is far less reliable than it is for
the low-expressing batch, just from having so few points.

```{r high-final-hits}
high$hit_results
```

```{r high-final-plot}
high$p_final
```

```{r high-save}
write_csv(high$hit_results, "../results/final_binder_ranking_high.csv")
```

## 5. Low-expressing batch

This is the batch that was getting compositionally squeezed in the unsplit
analysis - here it's normalised entirely on its own terms.

```{r low-batch-run}
low_batch_counts <- counts_long %>% filter(binder_id %in% low_expr_binders)
low <- run_batch_pipeline(low_batch_counts)
```

```{r low-innate-toxicity}
low$p_innate_toxicity
```

```{r low-key-comparison}
low$p_key_comparison
```

```{r low-final-hits}
low$hit_results
```

```{r low-final-plot}
low$p_final
```

```{r low-save}
write_csv(low$hit_results, "../results/final_binder_ranking_low.csv")
```

## 6. Comparing each batch to the published plate-based screen (Sen2 retron)

Same reference data and same caveats as `DMS_analysis.Rmd` Section 7 (their
effect size uses an IPTG-low baseline and a GFP correction we don't have -
this is a rough sanity check, not a strict validation) - run once per batch.

```{r load-plate-reference}
plate_reference <- read_tsv("../data/plate_assay_reference_hits.tsv", show_col_types = FALSE) %>%
  filter(retron == "Sen2") %>%
  transmute(genename, plate_pval = pval, plate_log2_efsize = log2(efsize), plate_hit = hit)
```

```{r plate-compare-batches}
cmp_high <- compare_to_plate(high$hit_results, plate_reference, "high-expressing")
cmp_low  <- compare_to_plate(low$hit_results,  plate_reference, "low-expressing")

print(cmp_high$plot)
print(cmp_low$plot)

bind_rows(cmp_high$cor_summary, cmp_low$cor_summary)
```

```{r save-plate-comparisons}
write_csv(cmp_high$comparison, "../results/plate_assay_comparison_high.csv")
write_csv(cmp_low$comparison, "../results/plate_assay_comparison_low.csv")
```

## 7. Did splitting help?

Compare the table above against `DMS_analysis.Rmd`'s unsplit
`plate_assay_comparison.csv` correlation. What to look for:

- If the low-expressing batch's correlation with the plate assay is
  noticeably better here than in the unsplit notebook, that's evidence the
  dominant clones really were distorting the comparison for everyone else -
  exactly the problem this notebook exists to fix.
- If it's about the same, the low-expressing binders' own read depth (not
  the dominant clones' compositional pull) is the limiting factor - check
  `n_excl_low_count` and how many of the low batch are flagged `low_count`;
  splitting can't fix a binder that was only ever backed by a handful of
  reads to begin with.
- The high-expressing batch's correlation is based on very few binders
  (see the caution in Section 4) - a single outlier among 5-10 points can
  swing that number a lot, so don't read much into it alone.

## Where to go from here

* If most binders land in one batch and almost none in the other, the
  `high_expr_threshold` in Section 2 is in the wrong place for this
  dataset - move it and re-knit; the histogram there is the thing to look
  at, not a fixed number.
* This splits on overall abundance once, globally. A binder that's
  middling in one condition but dominant in another would still be
  assigned to a single batch for every comparison - if that matters for
  your data, consider a per-condition rather than per-binder split.
* Everything in `DMS_analysis.Rmd`'s own "Where to go from here" still
  applies here too (the dose-response check, the missing GFP control, and
  what a second replicate would buy you).
